---
title: "Selecting K with OpTop"
subtitle: "Parametric tests and discrepancy indices for LDA models"
author: "Francesco Grossetti"
output:
  rmarkdown::html_vignette:
    toc: true
    math_method: katex
vignette: >
  %\VignetteIndexEntry{Selecting K with OpTop}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<style>
/* the html_vignette theme styles section h4s with a left margin and a bottom
   border; the author/date lines are h4s too and would inherit both */
h4.author, h4.date { margin-left: 0; border-bottom: none; }
/* wide display equations scroll horizontally instead of clipping */
.katex-display { overflow-x: auto; overflow-y: hidden; }
</style>

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.2,
  fig.align = "center"
)
library(OpTop)
library(data.table)
library(ggplot2)

# Precomputed results shipped with the package (a few KB): tables derived from
# the full fitted grid, plus the captured console output of the optimal_topic()
# calls below.
res <- readRDS(system.file("extdata", "optop-vignette-results.rds",
                           package = "OpTop"))
meta <- res$meta
idx_tab <- res$inaugural$index_table
word_r2 <- res$inaugural$word_snapshot
sim_chi <- res$sim$chi
sim_picks <- res$sim$picks
sim_idx_tab <- res$sim$index_table

# The fitted grid itself (~170 MB) cannot ship with the package and lives in
# data-raw/ on the maintainer's machine. When it is available, every
# optimal_topic() call below runs live; when it is not (e.g. R CMD check
# re-building the vignette from the pruned tarball), the same objects are
# assigned from the shipped results and the calls' captured console output is
# replayed, so the rendered document is the same either way.
models_path <- file.path("..", "data-raw", "VEM_models_inaugural.rds")
have_models <- file.exists(models_path)
if (have_models) {
  VEM_models <- readRDS(models_path)
  toks <- quanteda::data_corpus_inaugural |>
    quanteda::tokens(remove_punct = TRUE,
                     remove_symbols = TRUE,
                     remove_numbers = TRUE) |>
    quanteda::tokens_tolower() |>
    quanteda::tokens_remove(quanteda::stopwords())
  mydfm_sub <- quanteda::dfm_trim(quanteda::dfm(toks), min_termfreq = 5)
  weighted_dfm <- quanteda::dfm_weight(mydfm_sub, scheme = "prop")
} else {
  chi_seq <- res$inaugural$chi_by_q[["q0.95"]]
  chi_min <- data.table::copy(chi_seq)
  chi_by_q <- res$inaugural$chi_by_q
  chi_cal <- res$inaugural$chi_cal
  chi_cal_mm <- res$inaugural$chi_cal_mm
}
```

OpTop implements the parametric tests of Lewis and Grossetti (2022, JMLR
23(58)) for identifying the optimal number of topics of a Latent Dirichlet
Allocation (LDA) model, together with a family of regression-style
goodness-of-fit (GoF) indices. This vignette illustrates both on the corpus
of the paper's case study and concludes with a simulation study in which the
true number of topics is known.

The only step performed offline is the estimation of the LDA models
(`r length(meta$K_grid)` fits). Everything downstream, from every
`optimal_topic()` call to each table, figure, and number quoted in the text,
is computed from those fits.

## Setting up: corpus, grid, models

We use the U.S. Presidential Inaugural Address corpus distributed with
**quanteda**, the same corpus analyzed in the paper's case study, with a
standard preprocessing pipeline and a vocabulary trim. The grid spans
K = `r min(meta$K_grid)` to `r max(meta$K_grid)` in steps of 2.

```{r fitting, eval = FALSE}
library(OpTop)

ncores <- 6

# Tokenize
toks <- quanteda::data_corpus_inaugural |>
  quanteda::tokens(remove_punct = TRUE,
                   remove_symbols = TRUE,
                   remove_numbers = TRUE) |>
  quanteda::tokens_tolower() |>
  quanteda::tokens_remove(quanteda::stopwords())

# Create DFM (with raw counts as well as proportions)
mydfm        <- quanteda::dfm(toks)
mydfm_sub    <- quanteda::dfm_trim(mydfm, min_termfreq = 5)
weighted_dfm <- quanteda::dfm_weight(mydfm_sub, scheme = "prop")

K_grid <- seq(2, 200, by = 2)

# Estimate VEM models in parallel
VEM_models <- parallel::mclapply(
  K_grid,
  function(k) {
    message("Estimating LDA with K = ", k)
    topicmodels::LDA(x = mydfm_sub, k = k)
  },
  mc.cores = ncores
)
```

After trimming, the corpus has `r meta$ndoc` documents and
`r meta$nfeat_trimmed` features (down from `r meta$nfeat_full`). The VEM fits
above are unseeded, so re-running the pipeline may alter individual
statistics slightly; the numbers quoted in this vignette come from one such
run.

## The Test 1 statistic

The test evaluates overall model adequacy: how well a $K$-topic model
explains the corpus. It relies on the multinomial structure that LDA itself
assumes. A fitted $K$-topic model describes document $j$ by a vector of
document-topic weights $\theta_j^K$ and each topic by a distribution over
the $P$ vocabulary words, collected in the topic-word matrix $\phi^K$. The
model's fitted probability of word $p$ in document $j$ is

$$
I_{jp}^{K} = \sum_{k=1}^{K} \theta_{jk}^{K}\,\phi_{kp}^{K},
\tag{1}
$$

which is Equation 5 in the paper. Equation 4 of the paper represents the
observed word-proportion vector $d_j$ as this fitted distribution plus
multinomial sampling noise. If the $K$-topic model fully characterizes the
corpus, one would be unable to reject the hypothesis that the observed and
estimated word distributions are statistically indistinct: a classical
goodness-of-fit situation.

Because a large number of words carry near-zero fitted probability, the
comparison is made on a per-document **envelope**. The fitted probabilities
$I_{jp}^K$ are sorted in decreasing order and the smallest leading set whose
cumulative mass exceeds $q$ is retained; these are the $P_j$ relatively
important words of document $j$. The remaining, relatively unimportant words
are collapsed into a single bin (the paper's Equations 6 and 7). With the
words of document $j$ indexed in decreasing order of $I_{jp}^K$, the fitted
and observed masses of that bin are

$$
\begin{gathered}
I^{K}_{j,\min} = \sum_{p = P_j + 1}^{P} I^{K}_{jp} \;<\; 1 - q,
\\
D_{j,\min} = \sum_{p = P_j + 1}^{P} D_{jp}.
\end{gathered}
\tag{2}
$$

Collapsing the relatively unimportant words in this manner preserves the
underlying multinomial assumption. Note that the basis for the cutoff is the
fitted model rather than the observed documents; the paper's numerical
setting $I^K = 0.05$ corresponds to `q = 0.95`. Test 1 (Equation 8 of the
paper) then aggregates the Pearson discrepancies over documents:

$$
\begin{split}
\mathrm{OpTop}_J^K = \sum_{j=1}^{J} & \left[\, (P_j + 1) \left(
\sum_{p=1}^{P_j} \frac{\bigl(D_{jp} - I_{jp}^K\bigr)^2}{I_{jp}^K}
\right.\right. \\
& \left.\left. \quad
+\; \frac{\bigl(D_{j,\min} - I_{j,\min}^K\bigr)^2}{I_{j,\min}^K}
\right) \right] \sim \chi^2_{P_J},
\end{split}
\tag{3}
$$

where $\mathrm{OpTop}_J^K$ is distributed chi-square with
$P_J = \sum_{j=1}^{J} P_j$ degrees of freedom. Misspecification, that is a
$K$ too coarse to reproduce the word distributions, can only inflate the
statistic, so evidence against adequacy is found in the upper tail. The
package reports the **standardized** statistic $\mathrm{OpTop}_J^K / P_J$
(the version plotted in the paper's Figure 2, which fluctuates around 1
under an adequate model) in the `OpTop` column, the degrees of freedom $P_J$
in `df`, and the upper-tail p-value in `pval`.

## Running the test

`optimal_topic()` consumes **word proportions** (`weighted_dfm` above),
while the discrepancy indices of the second part consume the raw **counts**;
both objects are therefore required. With `verbose = TRUE`, the function
reports its setup, its progress across the grid, the selected model together
with the rule that chose it, and the elapsed time:

```{r optimal-call-seq, eval = have_models}
# Use the sequential approach
chi_seq <- optimal_topic(topic_models = VEM_models,
                         weighted_dfm = weighted_dfm,
                         selection = "sequential",
                         q = 0.95,
                         alpha = 0.01,
                         do_plot = FALSE)
```

```{r optimal-call-seq-replay, eval = !have_models, echo = FALSE}
cat(res$inaugural$console_seq, sep = "\n")
```

```{r optimal-call-min, eval = have_models}
# Use the global minimum approach
chi_min <- optimal_topic(topic_models = VEM_models,
                         weighted_dfm = weighted_dfm,
                         selection = "min",
                         q = 0.95,
                         alpha = 0.01,
                         do_plot = FALSE)
```

```{r optimal-call-min-replay, eval = !have_models, echo = FALSE}
cat(res$inaugural$console_min, sep = "\n")
```

## Reading the result

`optimal_topic()` returns a `data.table` with one row per model:

```{r table-head}
head(chi_seq)
```

- `OpTop` is the **standardized** Test 1 statistic, the raw chi-square of
  Equation (3) divided by its degrees of freedom, and the quantity plotted
  in the paper's Figure 2. Under an adequate model it fluctuates around 1.
- `df` is $P_J = \sum_j P_j$, the total number of relatively important words
  across documents.
- `pval` is the upper-tail p-value of the raw statistic. Small values
  indicate that the observed discrepancy is larger than sampling noise can
  explain, so the $K$-topic model is rejected.

The `selection` rule affects which K is *reported* as optimal (and the plot
guides), never the table itself. The two calls above return identical
tables:

```{r rule-independence}
all.equal(chi_seq, chi_min)
```

### The two selection rules

Both rules are direct computations on the returned table and can be stated
explicitly:

```{r selection-rules}
alpha <- 0.01

# "sequential": the smallest K the test fails to reject
seq_pick <- chi_seq[pval > alpha, topic][1]

# "min": the K with the minimum standardized statistic
min_pick <- chi_seq[which.min(OpTop), topic]

c(sequential = seq_pick, min = min_pick)
```

**`selection = "sequential"`** implements the classical sequential scan for
model order: $K$ is increased until the first model the test *fails to
reject* (`pval > alpha`). The rationale is parsimony: the rule selects the
smallest $K$ that the data cannot distinguish from adequate. On this corpus
the sequential rule selects **K = `r seq_pick`**. If every model were
rejected, `optimal_topic()` would fall back to the global minimum with a
warning.

**`selection = "min"`** reproduces the published case study: the optimum is
identified as the $K$-topic model with the minimum standardized statistic
across the grid. Here the minimum is at **K = `r min_pick`**, close to the
92 topics the paper identifies on the untrimmed vocabulary of the same
corpus and, as the sweep below shows, insensitive to the choice of `q`. One
practical remark is in order: the minimum rule requires a grid that extends
beyond the flattening of the curve. On a grid truncated at K = 100 the
minimum sits at the boundary, and only the full scan up to K = 200 allows it
to stabilize. The behavior has a structural explanation. The `df` column
decreases steadily with $K$ because the envelopes shrink as the fitted
distributions concentrate, and since the null level of the standardized
statistic scales with the envelope size, part of the curve's long decline
reflects envelope shrinkage rather than genuine improvement in fit. The
calibrated p-values of the next section absorb this effect automatically;
the raw curve does not. The sequential rule, by construction, stops at the
first adequate $K$ and is unaffected by the extent of the grid.

The p-value column explains when and why the two rules diverge:

```{r pval-summary}
summary(chi_seq$pval)
```

With `df` of the order of
`r format(round(median(chi_seq$df)), big.mark = ",")`, the chi-square
reference distribution is extremely concentrated, so p-values saturate:
models whose standardized statistic is materially above 1 receive p-values
of essentially 0, and models at or below 1 receive p-values of essentially
1. When every model is rejected, the sequential rule falls back to the
global minimum with a warning; when several models are accepted, it stops at
the smallest, which may precede the minimum by a wide margin. This is a
documented calibration caveat of the chi-square reference (see
`?optimal_topic`): in saturated regimes the information is carried by the
*shape* of the `OpTop` curve and by its minimum.

Because the return value is a table, custom figures follow directly:

```{r optop-curve}
ggplot(chi_seq, aes(x = topic, y = OpTop)) +
  geom_line(linewidth = 0.8, color = "royalblue") +
  geom_vline(xintercept = seq_pick, linetype = 2) +
  geom_vline(xintercept = min_pick, linetype = 3) +
  annotate("point", x = min_pick, y = min(chi_seq$OpTop),
           color = "red", shape = 4, size = 4) +
  labs(x = "Topics", y = expression(OpTop[J]^{K}),
       title = "Standardized Test 1 statistic across the grid",
       subtitle = sprintf("sequential pick: K = %d (dashed) | global minimum: K = %d (dotted)",
                          seq_pick, min_pick)) +
  theme_bw()
```

## Calibrated p-values

The saturation above has a precise cause: the $\chi^2_{P_J}$ reference is an
asymptotic guide rather than the exact null law of the statistic. Write
$k_j = P_j + 1$ for the number of envelope bins of document $j$ and $N_j$
for its length in tokens. The classical Pearson statistic on *counts*,
$X^2_j = N_j \sum_b (d_{jb} - p_{jb})^2 / p_{jb}$, is the object with the
textbook $\chi^2_{k_j - 1}$ asymptotics; the document term of Equation (3)
is instead

$$
S_j = (P_j + 1) \sum_b \frac{(d_{jb} - p_{jb})^2}{p_{jb}}
= \frac{k_j}{N_j}\, X^2_j ,
\tag{4}
$$

which differs from the calibrated object by the factor $k_j / N_j$. In
addition, the fitted probabilities are estimated from the same data they are
tested against. Calibration therefore replaces the $\chi^2$ reference with
the distribution of $T = \sum_j S_j$ under the **conditional fitted-model
null**

$$
c_j \sim \mathrm{Multinomial}\bigl(N_j,\; I_j^K\bigr),
\qquad j = 1, \dots, J,
\tag{5}
$$

with $\hat\theta$ and $\hat\phi$ held fixed at their estimates. Two
properties make this exact and fast: the envelope depends only on the model,
never on the data, and a multinomial collapsed over bins is multinomial on
the collapsed probabilities. The null replicates can therefore be drawn
*directly on the $k_j$ bins*
$p_j = (I^K_{j1}, \dots, I^K_{jP_j}, I^K_{j,\min})$, which is equivalent to
simulating entire documents over the vocabulary at a small fraction of the
cost. The only new ingredient the user supplies is the vector of document
lengths $N_j$:

```{r optimal-call-cal, eval = have_models}
doc_lens <- quanteda::ntoken(mydfm_sub)

chi_cal <- optimal_topic(topic_models = VEM_models,
                         weighted_dfm = weighted_dfm,
                         q = 0.95,
                         alpha = 0.01,
                         calibrate = "bootstrap",
                         n_boot = 200,
                         doc_lengths = doc_lens,
                         seed = 20260703,
                         do_plot = FALSE)
```

```{r optimal-call-cal-replay, eval = !have_models, echo = FALSE}
cat(res$inaugural$console_cal, sep = "\n")
```

With `calibrate = "bootstrap"`, $B$ = `n_boot` null replicates
$T^{*}_1, \dots, T^{*}_B$ are drawn as above and the reported p-value is the
empirical upper tail with the standard finite-sample correction,

$$
\hat{p} = \frac{1 + \#\{\, b : T^{*}_b \ge T \,\}}{B + 1},
\tag{6}
$$

so `alpha` is a genuine Type-I error rate with respect to the conditional
null, at resolution $1/(B+1)$. The table now carries two p-value columns:
`pval` (the calibrated value used by the selection rules) and `pval_chisq`
(the asymptotic value, kept for comparison):

```{r cal-table}
head(chi_cal)
```

The bootstrap is the reference method; `calibrate = "moment"` targets the
same null in closed form. Haldane (1937) gives the exact multinomial moments
of the count-based Pearson statistic over $k_j$ bins:

$$
\begin{split}
\mathrm{E}\bigl[X^2_j\bigr] &= k_j - 1,
\\
\mathrm{Var}\bigl[X^2_j\bigr] &= 2(k_j - 1)
\\
&\quad + \frac{\sum_b 1/p_{jb} - k_j^2 - 2k_j + 2}{N_j}.
\end{split}
\tag{7}
$$

Since $S_j = (k_j/N_j)\,X^2_j$ and documents are independent, the null mean
and variance of $T$ follow by scaling and summing,
$\mu = \sum_j (k_j/N_j)(k_j - 1)$ and
$\sigma^2 = \sum_j (k_j/N_j)^2\,\mathrm{Var}[X^2_j]$, and the reference is
the Satterthwaite scaled chi-square matching them:

$$
\begin{gathered}
T \overset{\text{approx}}{\sim} a\,\chi^2_\nu,
\qquad a = \frac{\sigma^2}{2\mu},
\qquad \nu = \frac{2\mu^2}{\sigma^2},
\\
\hat{p} = \Pr\!\left(\chi^2_\nu \ge T/a\right).
\end{gathered}
\tag{8}
$$

The method requires no simulation and no seed; it is a fast approximation
that corrects the location and the scale of the reference, although not its
higher moments:

```{r optimal-call-mm, eval = have_models}
chi_cal_mm <- optimal_topic(topic_models = VEM_models,
                            weighted_dfm = weighted_dfm,
                            q = 0.95,
                            alpha = 0.01,
                            calibrate = "moment",
                            doc_lengths = doc_lens,
                            do_plot = FALSE)
```

```{r optimal-call-mm-replay, eval = !have_models, echo = FALSE}
cat(res$inaugural$console_cal_mm, sep = "\n")
```

The following figure compares the three reference distributions:

```{r cal-plot}
pv <- melt(
  data.table(topic = chi_cal$topic,
             `chi-square` = chi_cal$pval_chisq,
             bootstrap = chi_cal$pval,
             moment = chi_cal_mm$pval),
  id.vars = "topic", variable.name = "reference", value.name = "pvalue"
)

ggplot(pv, aes(x = topic, y = pvalue, color = reference)) +
  geom_line(linewidth = 0.7) +
  geom_hline(yintercept = alpha, linetype = 3) +
  labs(x = "Topics", y = "p-value",
       title = "P-values under the three reference distributions",
       subtitle = sprintf("dotted line: alpha = %.2f", alpha)) +
  theme_bw()
```

Because the p-values themselves do not depend on `alpha`, the conventional
thresholds can be scanned directly on the tables already computed:

```{r cal-alpha-sweep}
alphas <- c(0.01, 0.05, 0.10)

picks_alpha <- data.table(
  alpha = alphas,
  bootstrap = sapply(alphas, function(a) chi_cal[pval > a, topic][1]),
  moment = sapply(alphas, function(a) chi_cal_mm[pval > a, topic][1]),
  `chi-square` = sapply(alphas, function(a) chi_seq[pval > a, topic][1])
)

knitr::kable(picks_alpha,
             caption = "Sequential pick by reference distribution as alpha varies")
```

Interpretation follows from the null the calibration targets. The calibrated
sequential pick answers a precise question: *what is the smallest $K$ whose
residual discrepancy could plausibly be produced by multinomial sampling
noise alone, if the fitted model were the truth?* On this corpus this
happens already at K = `r chi_cal[pval > alpha, topic][1]`. The fit does
continue to improve beyond that point, since the `OpTop` curve decreases
until its minimum at K = `r min_pick`, but from there onward the
improvements are smaller than what sampling noise can account for at the
chosen `alpha`.

The uncalibrated scan is best understood as a *reduced-chi-square
heuristic*. With `df` this large the asymptotic p-values saturate at 0 or 1,
so in practice the rule stops at the first $K$ whose standardized
discrepancy falls below 1. This is a reasonable descriptive criterion, but
its implicit threshold is not tied to the actual null level of the
statistic. On this corpus the bootstrap places that null level closer to 2
than to 1, so the heuristic demands more than noise-level adequacy and stops
later, at K = `r seq_pick`. The selection is also unstable: on this grid the
next model, K = `r seq_pick + 2`, is rejected again, and the sweep of `q`
below moves the pick from 20 to 38. When the implicit threshold happens to
lie close to the true null level, the same heuristic performs well; the
simulation study at the end of this vignette, where the ground truth is
known, is such a case. What calibration adds is therefore not a different
answer but a guarantee. Raising `alpha` demands stronger evidence of
adequacy, so the pick moves weakly upward, as the table shows, and `alpha`
becomes a genuine Type-I error rate with respect to the fitted-model null,
at bootstrap resolution `1 / (n_boot + 1)`. In summary, the calibrated
sequential rule identifies the most parsimonious *statistically adequate*
$K$, while the global-minimum rule and the discrepancy indices below
quantify *maximal descriptive fit*. The two questions are distinct.

One caveat applies to both calibrations: they hold the estimated `theta` and
`phi` fixed, so they do not account for estimation noise in the LDA
parameters themselves (see `?optimal_topic`).

## The role of `q`

`q` controls the per-document envelope introduced in Equation (2): the
fitted word probabilities are sorted in decreasing order and the smallest
leading set whose cumulative mass exceeds `q` is retained, while the
relatively unimportant words are collapsed into a single bin whose mass
remains strictly below `1 - q`. In the paper's notation `q = 1 - I^K`, and
its numerical setting `I^K = 0.05` corresponds to the default `q = 0.95`.

Re-running the test across a sweep of `q` values is a direct sensitivity
check:

```{r q-sweep-run, eval = have_models}
q_grid <- c(0.80, 0.90, 0.95, 0.99)

chi_by_q <- lapply(q_grid, function(q) {
  optimal_topic(VEM_models, weighted_dfm, q = q, alpha = alpha,
                do_plot = FALSE)
})
names(chi_by_q) <- sprintf("q%.2f", q_grid)
```

```{r q-sweep-plot}
q_curves <- rbindlist(
  lapply(names(chi_by_q), function(nm) {
    data.table(q = as.numeric(sub("q", "", nm)), chi_by_q[[nm]])
  })
)

ggplot(q_curves, aes(x = topic, y = OpTop, color = factor(q))) +
  geom_line(linewidth = 0.7) +
  labs(x = "Topics", y = expression(OpTop[J]^{K}), color = "q",
       title = "Standardized Test 1 statistic across the q sweep") +
  theme_bw()
```

Applying the two selection rules to each element of the sweep shows how the
selected $K$ responds to `q`:

```{r q-picks}
picks_by_q <- data.table(
  q = as.numeric(sub("q", "", names(chi_by_q))),
  sequential = sapply(chi_by_q, function(tab) tab[pval > alpha, topic][1]),
  min = sapply(chi_by_q, function(tab) tab[which.min(OpTop), topic])
)

knitr::kable(picks_by_q, caption = "Selected K by rule as q varies")
```

Larger `q` retains more words per document, so the degrees of freedom grow
and the envelope covers more of each document's distribution, while smaller
`q` makes the comparison coarser and computationally cheaper. The sequential
pick moves with `q`, whereas the global minimum is unaffected. The default
`q = 0.95` follows the paper.

## Goodness-of-fit: the discrepancy indices

The second family of tools asks a complementary question: not *which K is
optimal* but *how much of the word-count structure a given model explains*.
The indices compare each model against the no-topics corpus baseline on a
**harmonized support**: a fixed, document-specific set of rare words
(computed across the whole grid) is collapsed so that every K is evaluated
on the same bins. They consume raw counts:

```{r indices-call, eval = FALSE}
dtm_counts <- methods::as(mydfm_sub, "CsparseMatrix")
partition <- optop_make_partition(VEM_models, dtm_counts, c = 5)
baseline <- optop_make_baseline(dtm_counts)

idx_tab <- optop_index_table(VEM_models, dtm_counts,
                             metrics = c("se", "chisq", "deviance"),
                             partition = partition, baseline = baseline,
                             macro = TRUE)
```

```{r indices-plot}
idx <- melt(idx_tab,
            id.vars = "K",
            measure.vars = c("R2_SE", "R2_chisq", "R2_dev"),
            variable.name = "metric", value.name = "R2")

ggplot(idx, aes(x = K, y = R2, color = metric)) +
  geom_line(linewidth = 0.7) +
  labs(x = "Topics", y = expression(R^2),
       title = "Discrepancy indices across the grid (document level, micro)") +
  theme_bw()
```

All three indices are regression-style R-squared analogues: 0 indicates that
the model explains no more than the corpus-wide baseline, and 1 indicates a
perfect account of the harmonized word counts. They typically rise steeply
at small $K$ and flatten once additional topics no longer provide
significant incremental explanatory power; the region where the curves
flatten corresponds to the neighborhood in which Test 1 locates the optimum.

The indices are also available at the **word level**, which serves as a
diagnostic: it identifies the words the selected model captures well and
those it misses. For the sequential pick (K = `r word_r2$k`), the micro
word-level chi-square index is `r round(word_r2$micro, 3)`; the ten highest-
and lowest-index words are reported below:

```{r word-snapshot}
knitr::kable(
  data.frame(
    `best words` = names(word_r2$best),
    R2_best = round(unname(word_r2$best), 3),
    `worst words` = names(word_r2$worst),
    R2_worst = round(unname(word_r2$worst), 3),
    check.names = FALSE
  ),
  caption = sprintf("Word-level chi-square R2 at K = %d: the ten best and worst captured words",
                    word_r2$k)
)
```

## A simulation with known truth

In empirical corpora the true number of topics is unknown. Following the
paper, we therefore validate the procedure through a simulation study in
which the ground truth is known: one corpus is simulated from the LDA
generative process itself, with `r meta$J_sim` documents over
`r meta$W_sim` words drawn from **`r meta$true_K` topics**, using
`sim_dfm()` (a thin wrapper around `LDATS::sim_LDA_data()`), and a grid of
models is re-estimated around the truth.

```{r sim-call, eval = FALSE}
rdirichlet <- function(n, alpha) {
  g <- matrix(rgamma(n * length(alpha), shape = alpha, rate = 1),
              nrow = n, byrow = TRUE)
  g / rowSums(g)
}

set.seed(20260703)
true_K <- 10L
Theta_true <- rdirichlet(100L, rep(0.3, true_K))    # documents x topics
Beta_true <- rdirichlet(true_K, rep(0.05, 1500L))   # topics x words
doc_len <- sample(500:1500, 100L, replace = TRUE)

sim_corpus <- sim_dfm(DTW = Theta_true, TWW = Beta_true,
                      doc_length = doc_len, seed = 42)

sim_models <- parallel::mclapply(
  seq(2, 20, by = 2),
  function(k) topicmodels::LDA(sim_corpus, k = k,
                               control = list(seed = 2000 + k)),
  mc.cores = ncores
)

sim_chi <- optimal_topic(sim_models,
                         quanteda::dfm_weight(sim_corpus, scheme = "prop"),
                         q = 0.95, alpha = alpha, verbose = TRUE)
```

```{r sim-plot}
ggplot(sim_chi, aes(x = topic, y = OpTop)) +
  geom_line(linewidth = 0.8, color = "royalblue") +
  geom_vline(xintercept = meta$true_K, color = "red", linetype = 1) +
  geom_vline(xintercept = sim_picks$sequential, linetype = 2) +
  geom_vline(xintercept = sim_picks$min, linetype = 3) +
  labs(x = "Topics", y = expression(OpTop[J]^{K}),
       title = "Simulated corpus with ground truth K = 10 (solid red)",
       subtitle = sprintf("sequential pick: K = %d (dashed) | global minimum: K = %d (dotted)",
                          sim_picks$sequential, sim_picks$min)) +
  theme_bw()
```

```{r sim-indices}
sim_idx <- melt(sim_idx_tab,
                id.vars = "K",
                measure.vars = c("R2_SE", "R2_chisq", "R2_dev"),
                variable.name = "metric", value.name = "R2")

ggplot(sim_idx, aes(x = K, y = R2, color = metric)) +
  geom_line(linewidth = 0.7) +
  geom_vline(xintercept = meta$true_K, color = "red", linetype = 1) +
  labs(x = "Topics", y = expression(R^2),
       title = "Discrepancy indices on the simulated corpus") +
  theme_bw()
```

The statistic bends in the neighborhood of the true $K$ and the indices stop
improving there, the same qualitative behavior reported in the paper's
simulation study (its Figure 7). As the paper acknowledges, the identified
optimum is not always exactly $K_{sim}$ but is typically adjacent or in
close proximity to it: here the sequential rule stops at
K = `r sim_picks$sequential` with the truth at `r meta$true_K`, while the
global minimum moves toward the grid boundary as the curve continues to
decrease past the bend. Note that this is the *uncalibrated* sequential rule
performing well. On this corpus the post-truth plateau of the standardized
statistic lies below 1, so the implicit reduced-chi-square threshold of the
saturated p-values falls close to the actual null level, which is the
favorable case discussed in the calibration section. The tools should be
read as locating the *region* of adequate $K$, with the discrepancy indices
quantifying the incremental explanatory power of each candidate.

## Notes

- The legacy stability helpers (`topic_stability()`, `agg_topic_stability()`,
  `agg_document_stability()`, `get_topic_models()`) are deprecated and will
  be removed in a future release; the discrepancy indices shown here are
  their replacement.
- `selection = "legacy"` reproduces the pre-0.9.9 computation exactly for
  back-compatibility and is also deprecated.
- The chi-square reference for Test 1 is asymptotic rather than exact; the
  calibrated p-values presented above (`calibrate = "bootstrap"` or
  `calibrate = "moment"`) address this, and `?optimal_topic` documents the
  underlying statistical reasoning.

## References

Haldane, J. B. S. (1937). The exact value of the moments of the distribution
of chi-square. *Biometrika*, 29, 133–143.

Lewis, C. M. and Grossetti, F. (2022). A statistical approach for optimal
topic model identification. *Journal of Machine Learning Research*, 23(58),
1–20. <https://jmlr.org/papers/v23/19-297.html>
