---
title: "Selecting K with OpTop"
subtitle: "Parametric tests and discrepancy indices for LDA models"
author: "Francesco Grossetti"
output:
  rmarkdown::html_vignette:
    toc: true
    math_method: katex
vignette: >
  %\VignetteIndexEntry{Selecting K with OpTop}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<style>
/* the html_vignette theme styles section h4s with a left margin and a bottom
   border; the author/date lines are h4s too and would inherit both */
h4.author, h4.date { margin-left: 0; border-bottom: none; }
</style>

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.2,
  fig.align = "center"
)
library(OpTop)
library(data.table)
library(ggplot2)

# Precomputed results shipped with the package (a few KB): tables derived from
# the full fitted grid, plus the captured console output of the optimal_topic()
# calls below.
res <- readRDS(system.file("extdata", "optop-vignette-results.rds",
                           package = "OpTop"))
meta <- res$meta
idx_tab <- res$inaugural$index_table
word_r2 <- res$inaugural$word_snapshot
sim_chi <- res$sim$chi
sim_picks <- res$sim$picks
sim_idx_tab <- res$sim$index_table

# The fitted grid itself (~170 MB) cannot ship with the package and lives in
# data-raw/ on the maintainer's machine. When it is available, every
# optimal_topic() call below runs live; when it is not (e.g. R CMD check
# re-building the vignette from the pruned tarball), the same objects are
# assigned from the shipped results and the calls' captured console output is
# replayed, so the rendered document is the same either way.
models_path <- file.path("..", "data-raw", "VEM_models_inaugural.rds")
have_models <- file.exists(models_path)
if (have_models) {
  VEM_models <- readRDS(models_path)
  toks <- quanteda::data_corpus_inaugural |>
    quanteda::tokens(remove_punct = TRUE,
                     remove_symbols = TRUE,
                     remove_numbers = TRUE) |>
    quanteda::tokens_tolower() |>
    quanteda::tokens_remove(quanteda::stopwords())
  mydfm_sub <- quanteda::dfm_trim(quanteda::dfm(toks), min_termfreq = 5)
  weighted_dfm <- quanteda::dfm_weight(mydfm_sub, scheme = "prop")
} else {
  chi_seq <- res$inaugural$chi_by_q[["q0.95"]]
  chi_min <- data.table::copy(chi_seq)
  chi_by_q <- res$inaugural$chi_by_q
  chi_cal <- res$inaugural$chi_cal
  chi_cal_mm <- res$inaugural$chi_cal_mm
}
```

OpTop implements the parametric tests of Lewis and Grossetti (2022, JMLR
23(58)) for identifying the optimal number of topics of a Latent Dirichlet
Allocation (LDA) model, together with a family of regression-style
goodness-of-fit indices. This vignette walks through both on a real corpus,
then closes with a simulation where the true number of topics is known.

The only step run offline is the LDA fitting itself
(`r length(meta$K_grid)` models); everything downstream — every
`optimal_topic()` call, table, plot, and number quoted in the text — is
computed from those fits.

## Setting up: corpus, grid, models

We use the U.S. Presidential Inaugural Address corpus that ships with
**quanteda** — the same corpus as the paper's case study — with a standard
preprocessing pipeline and a vocabulary trim. The grid spans
K = `r min(meta$K_grid)` to `r max(meta$K_grid)` in steps of 2.

```{r fitting, eval = FALSE}
library(OpTop)

ncores <- 6

# Tokenize
toks <- quanteda::data_corpus_inaugural |>
  quanteda::tokens(remove_punct = TRUE,
                   remove_symbols = TRUE,
                   remove_numbers = TRUE) |>
  quanteda::tokens_tolower() |>
  quanteda::tokens_remove(quanteda::stopwords())

# Create DFM (with raw counts as well as proportions)
mydfm        <- quanteda::dfm(toks)
mydfm_sub    <- quanteda::dfm_trim(mydfm, min_termfreq = 5)
weighted_dfm <- quanteda::dfm_weight(mydfm_sub, scheme = "prop")

K_grid <- seq(2, 200, by = 2)

# Estimate VEM models in parallel
VEM_models <- parallel::mclapply(
  K_grid,
  function(k) {
    message("Estimating LDA with K = ", k)
    topicmodels::LDA(x = mydfm_sub, k = k)
  },
  mc.cores = ncores
)
```

After trimming, the corpus has `r meta$ndoc` documents and
`r meta$nfeat_trimmed` features (down from `r meta$nfeat_full`). The VEM fits
above are unseeded, so re-running the pipeline can shift individual
statistics slightly; the numbers quoted in this vignette come from one such
run.

## The Test 1 statistic

The test rests on the multinomial structure that LDA itself assumes. A fitted
$K$-topic model describes document $j$ by a vector of document–topic weights
$\theta_j^K$ and every topic by a distribution over the $P$ vocabulary words,
collected in the topic–word matrix $\phi^K$. The model's fitted probability
of word $p$ in document $j$ is therefore

$$
I_{jp}^{K} \;=\; \sum_{k=1}^{K} \theta_{jk}^{K}\,\phi_{kp}^{K},
$$

and the paper's Equation 4 views the observed word-proportion vector $d_j$ as
this fitted distribution plus multinomial sampling noise. If the $K$-topic
model fully characterizes the corpus, the observed and fitted word
distributions should be statistically indistinguishable — a classical
goodness-of-fit situation.

Because most of the vocabulary carries near-zero fitted probability, the
comparison is made on a per-document **envelope**: sort $I_{jp}^K$ in
decreasing order and keep the smallest head whose cumulative mass exceeds
$q$ — these are the $P_j$ "relatively important" words — while the remaining
words are collapsed into a single bin with fitted mass
$I^K_{j,\min} = \sum_{p \notin \text{head}} I^K_{jp} < 1 - q$ and observed
mass $D_{j,\min}$ (Equations 5–7; the paper's numerical setting
$I^K = 0.05$ corresponds to `q = 0.95`). Test 1 (Equation 8) then aggregates
the Pearson discrepancies over documents:

$$
\mathrm{OpTop}_J^K \;=\; \sum_{j=1}^{J} \left[\, (P_j + 1) \left(
\sum_{p=1}^{P_j} \frac{\bigl(D_{jp} - I_{jp}^K\bigr)^2}{I_{jp}^K}
\;+\; \frac{\bigl(D_{j,\min} - I_{j,\min}^K\bigr)^2}{I_{j,\min}^K}
\right) \right]
\;\sim\; \chi^2_{P_J},
\qquad P_J = \sum_{j=1}^{J} P_j .
$$

Misspecification — a $K$ too coarse to reproduce the word distributions —
can only inflate the statistic, so evidence against adequacy lives in the
upper tail. The package reports the **standardized** statistic
$\mathrm{OpTop}_J^K / P_J$ (the version plotted in the paper's Figure 2,
fluctuating around 1 under a model that fits) in the `OpTop` column, the
degrees of freedom $P_J$ in `df`, and the upper-tail p-value in `pval`.

## Running the test

`optimal_topic()` consumes **word proportions** (`weighted_dfm` above), while
the discrepancy indices of the second part consume the raw **counts** — keep
both objects around. With `verbose = TRUE` the function reports its setup,
progress across the grid, the selected model with the rule that chose it, and
the elapsed time:

```{r optimal-call-seq, eval = have_models}
# Use the sequential approach
chi_seq <- optimal_topic(topic_models = VEM_models,
                         weighted_dfm = weighted_dfm,
                         selection = "sequential",
                         q = 0.95,
                         alpha = 0.01,
                         do_plot = FALSE)
```

```{r optimal-call-seq-replay, eval = !have_models, echo = FALSE}
cat(res$inaugural$console_seq, sep = "\n")
```

```{r optimal-call-min, eval = have_models}
# Use the global minimum approach
chi_min <- optimal_topic(topic_models = VEM_models,
                         weighted_dfm = weighted_dfm,
                         selection = "min",
                         q = 0.95,
                         alpha = 0.01,
                         do_plot = FALSE)
```

```{r optimal-call-min-replay, eval = !have_models, echo = FALSE}
cat(res$inaugural$console_min, sep = "\n")
```

## Reading the result

`optimal_topic()` returns a `data.table` with one row per model:

```{r table-head}
head(chi_seq)
```

- `OpTop` is the **standardized** Test 1 statistic (the raw chi-square of
  Equation 8 divided by its degrees of freedom) — the quantity plotted in the
  paper's Figure 2. Under a model that fits, it fluctuates around 1.
- `df` is `sum_j P_j`, the total count of "relatively important" words across
  documents.
- `pval` is the upper-tail p-value of the raw statistic: small values mean
  the observed discrepancy is larger than sampling noise can explain, i.e.
  the K-topic model is rejected.

The `selection` rule affects which K is *reported* as optimal (and the plot
guides), never the table itself — the two calls above return the same data:

```{r rule-independence}
all.equal(chi_seq, chi_min)
```

### The two selection rules

Both rules are one-liners on the returned table, so it is worth seeing them
spelled out:

```{r selection-rules}
alpha <- 0.01

# "sequential": the smallest K the test fails to reject
seq_pick <- chi_seq[pval > alpha, topic][1]

# "min": the K with the minimum standardized statistic
min_pick <- chi_seq[which.min(OpTop), topic]

c(sequential = seq_pick, min = min_pick)
```

**`selection = "sequential"`** implements the classical scan for model order:
walk K upward and stop at the first model the test *fails to reject*
(`pval > alpha`). The idea is parsimony — take the smallest K that the data
cannot distinguish from adequate. On this corpus the sequential rule selects
**K = `r seq_pick`**. (If every model were rejected, `optimal_topic()` falls
back to the global minimum with a warning.)

**`selection = "min"`** reproduces the published case study: take the K with
the minimum standardized statistic. Here that is **K = `r min_pick`** —
remarkably close to the 92 the paper reports on the untrimmed vocabulary of
the same corpus, and (as the q-sweep below shows) insensitive to the choice
of `q`. One practical remark: the minimum rule needs a grid that extends past
the flattening of the curve to settle — on a grid truncated at K = 100 the
minimum sits at the boundary, and only the full K ≤ 200 scan lets it come to
rest. There is a structural reason for that creep: the `df` column falls
steadily with K (the envelopes shrink as the fitted distributions
concentrate), and since the null level of the standardized statistic scales
with the envelope size, part of the curve's long decline reflects envelope
shrinkage rather than genuine fit improvement. The calibrated p-values of the
next section absorb this drift automatically; the raw curve does not. The
sequential rule, by construction, stops at the first adequate K and does not
care how far the grid extends.

A look at the p-value column explains when and why the two rules diverge:

```{r pval-summary}
summary(chi_seq$pval)
```

With `df` of the order of
`r format(round(median(chi_seq$df)), big.mark = ",")`, the chi-square
reference distribution is extremely concentrated, so p-values saturate:
models whose standardized statistic is materially above 1 get p-values of
essentially 0, and models at or below 1 get p-values of essentially 1. When
every model is rejected, the sequential rule falls back to the global minimum
(with a warning); when several are accepted it stops at the smallest — which
may sit well before the minimum. This is a documented calibration caveat of
the chi-square yardstick (see `?optimal_topic`): in saturated regimes the
*shape* of the `OpTop` curve, and its minimum, carry the information.

Because the table is the return value, any plot is one `ggplot2` call away:

```{r optop-curve}
ggplot(chi_seq, aes(x = topic, y = OpTop)) +
  geom_line(linewidth = 0.8, color = "royalblue") +
  geom_vline(xintercept = seq_pick, linetype = 2) +
  geom_vline(xintercept = min_pick, linetype = 3) +
  annotate("point", x = min_pick, y = min(chi_seq$OpTop),
           color = "red", shape = 4, size = 4) +
  labs(x = "Topics", y = expression(OpTop[J]^{K}),
       title = "Standardized Test 1 statistic across the grid",
       subtitle = sprintf("sequential pick: K = %d (dashed) | global minimum: K = %d (dotted)",
                          seq_pick, min_pick)) +
  theme_bw()
```

## Calibrated p-values

The saturation above has a precise cause: the $\chi^2_{P_J}$ reference is a
yardstick, not the exact null law of the statistic. Write $k_j = P_j + 1$ for
the number of envelope bins of document $j$ and $N_j$ for its length in
tokens. The classical Pearson statistic on *counts*,
$X^2_j = N_j \sum_b (d_{jb} - p_{jb})^2 / p_{jb}$, is the object with the
textbook $\chi^2_{k_j - 1}$ asymptotics; the document term of Equation 8 is
instead

$$
S_j \;=\; (P_j + 1) \sum_b \frac{(d_{jb} - p_{jb})^2}{p_{jb}}
\;=\; \frac{k_j}{N_j}\, X^2_j ,
$$

off from the calibrated object by the factor $k_j / N_j$ — and on top of
that, the fitted probabilities are estimated from the same data they are
tested against. Calibration therefore replaces the $\chi^2$ reference with
the distribution of $T = \sum_j S_j$ under the **conditional fitted-model
null**

$$
c_j \;\sim\; \mathrm{Multinomial}\bigl(N_j,\; I_j^K\bigr),
\qquad j = 1, \dots, J,
$$

with $\hat\theta$ and $\hat\phi$ held fixed at their estimates. Two
properties make this exact and fast: the envelope depends only on the model
(never on the data), and a multinomial collapsed over bins is multinomial on
the collapsed probabilities — so the null replicates can be drawn *directly
on the $k_j$ bins* $p_j = (I^K_{j1}, \dots, I^K_{jP_j}, I^K_{j,\min})$,
which is equivalent to simulating whole documents over the vocabulary at a
tiny fraction of the cost. The only new ingredient the user supplies is the
vector of document lengths $N_j$:

```{r optimal-call-cal, eval = have_models}
doc_lens <- quanteda::ntoken(mydfm_sub)

chi_cal <- optimal_topic(topic_models = VEM_models,
                         weighted_dfm = weighted_dfm,
                         q = 0.95,
                         alpha = 0.01,
                         calibrate = "bootstrap",
                         n_boot = 200,
                         doc_lengths = doc_lens,
                         seed = 20260703,
                         do_plot = FALSE)
```

```{r optimal-call-cal-replay, eval = !have_models, echo = FALSE}
cat(res$inaugural$console_cal, sep = "\n")
```

With `calibrate = "bootstrap"`, $B$ = `n_boot` null replicates
$T^{*}_1, \dots, T^{*}_B$ are drawn as above and the reported p-value is the
empirical upper tail with the standard finite-sample correction,

$$
\hat{p} \;=\; \frac{1 + \#\{\, b : T^{*}_b \ge T \,\}}{B + 1},
$$

so `alpha` is a genuine Type-I error rate with respect to the conditional
null, at resolution $1/(B+1)$. The bootstrap is the most expensive stage of
the pipeline; its cost, and the effect of the `n_threads` argument, are
quantified in the section [A Note On Computational Efficiency]. The table
now carries two p-value columns:
`pval` (calibrated — what the selection rules use) and `pval_chisq` (the
asymptotic value, kept for comparison):

```{r cal-table}
head(chi_cal)
```

The bootstrap is the reference method; `calibrate = "moment"` targets the
same null in closed form. Haldane (1937) gives the exact multinomial moments
of the count-based Pearson statistic over $k_j$ bins:

$$
\mathrm{E}\bigl[X^2_j\bigr] = k_j - 1,
\qquad
\mathrm{Var}\bigl[X^2_j\bigr] = 2(k_j - 1) +
\frac{\sum_b 1/p_{jb} - k_j^2 - 2k_j + 2}{N_j}.
$$

Since $S_j = (k_j/N_j)\,X^2_j$ and documents are independent, the null mean
and variance of $T$ follow by scaling and summing,
$\mu = \sum_j (k_j/N_j)(k_j - 1)$ and
$\sigma^2 = \sum_j (k_j/N_j)^2\,\mathrm{Var}[X^2_j]$, and the reference is
the Satterthwaite scaled chi-square matching them:

$$
T \;\overset{\text{approx}}{\sim}\; a\,\chi^2_\nu,
\qquad a = \frac{\sigma^2}{2\mu},
\qquad \nu = \frac{2\mu^2}{\sigma^2},
\qquad \hat{p} = \Pr\!\left(\chi^2_\nu \ge T/a\right).
$$

No simulation, no seed — a fast approximation that corrects the location and
scale of the reference but not its higher moments:

```{r optimal-call-mm, eval = have_models}
chi_cal_mm <- optimal_topic(topic_models = VEM_models,
                            weighted_dfm = weighted_dfm,
                            q = 0.95,
                            alpha = 0.01,
                            calibrate = "moment",
                            doc_lengths = doc_lens,
                            do_plot = FALSE)
```

```{r optimal-call-mm-replay, eval = !have_models, echo = FALSE}
cat(res$inaugural$console_cal_mm, sep = "\n")
```

The three references side by side make the point visually:

```{r cal-plot}
pv <- melt(
  data.table(topic = chi_cal$topic,
             `chi-square` = chi_cal$pval_chisq,
             bootstrap = chi_cal$pval,
             moment = chi_cal_mm$pval),
  id.vars = "topic", variable.name = "reference", value.name = "pvalue"
)

ggplot(pv, aes(x = topic, y = pvalue, color = reference)) +
  geom_line(linewidth = 0.7) +
  geom_hline(yintercept = alpha, linetype = 3) +
  labs(x = "Topics", y = "p-value",
       title = "Calibration replaces a saturated reference with a usable one",
       subtitle = sprintf("dotted line: alpha = %.2f", alpha)) +
  theme_bw()
```

Because the p-values themselves do not depend on `alpha`, scanning the
conventional thresholds is a one-liner per method on the tables we already
have:

```{r cal-alpha-sweep}
alphas <- c(0.01, 0.05, 0.10)

picks_alpha <- data.table(
  alpha = alphas,
  bootstrap = sapply(alphas, function(a) chi_cal[pval > a, topic][1]),
  moment = sapply(alphas, function(a) chi_cal_mm[pval > a, topic][1]),
  `chi-square` = sapply(alphas, function(a) chi_seq[pval > a, topic][1])
)

knitr::kable(picks_alpha,
             caption = "Sequential pick by reference distribution as alpha varies")
```

How should this be read? The calibrated sequential pick answers a precise
question: *what is the smallest K whose leftover discrepancy could plausibly
be produced by multinomial sampling noise alone, if the fitted model were the
truth?* On this corpus that happens already at
K = `r chi_cal[pval > alpha, topic][1]`. The fit does keep improving beyond
that — the `OpTop` curve keeps falling until its minimum at
K = `r min_pick` — but from there onward the improvements are smaller than
what sampling noise can account for at the chosen `alpha`.

The uncalibrated scan is best understood as a *reduced-chi-square heuristic*:
with `df` this large the asymptotic p-values saturate to 0 or 1, so in
practice the rule stops at the first K where the standardized discrepancy
drops below 1 — a sensible engineering criterion, but one whose implicit
threshold is not tied to the actual null level of the statistic. On this
corpus the bootstrap places that null level nearer 2 than 1, so the heuristic
demands more than noise-level adequacy and stops later, at
K = `r seq_pick` — and fragilely so: on this grid the very next model,
K = `r seq_pick + 2`, is rejected again, and the q-sweep below moves the pick
from 20 to 38. When the implicit threshold happens to sit close to the true
null level the same heuristic performs well — the simulation at the end of
this vignette, where the truth is known, is exactly such a case. What
calibration adds is therefore not a different answer but a guarantee: raising
`alpha` demands stronger evidence of adequacy, so the pick moves (weakly)
upward, as the table shows — and `alpha` is now a real Type-I error rate with
respect to the fitted-model null (at bootstrap resolution
`1 / (n_boot + 1)`). In short: the calibrated sequential rule identifies the
most parsimonious *statistically adequate* K, while the global-minimum rule
and the discrepancy indices below quantify *maximal descriptive fit* — two
different questions, and it is worth knowing which one you are asking.

One caveat applies to both calibrations: they hold the estimated `theta` and
`phi` fixed, so they do not account for estimation noise in the LDA
parameters themselves (see `?optimal_topic`).

## The role of `q`

`q` controls the per-document envelope: sort each document's fitted word
probabilities in decreasing order and keep the smallest head whose cumulative
mass exceeds `q` — those are the `P_j` "relatively important" words; the rest
are collapsed into a single bin whose mass stays strictly below `1 - q`. In
the paper's notation `q = 1 - I^K`; its numerical setting `I^K = 0.05`
corresponds to the default `q = 0.95`.

Re-running the test across a sweep of `q` values is a direct sensitivity
check:

```{r q-sweep-run, eval = have_models}
q_grid <- c(0.80, 0.90, 0.95, 0.99)

chi_by_q <- lapply(q_grid, function(q) {
  optimal_topic(VEM_models, weighted_dfm, q = q, alpha = alpha,
                do_plot = FALSE)
})
names(chi_by_q) <- sprintf("q%.2f", q_grid)
```

```{r q-sweep-plot}
q_curves <- rbindlist(
  lapply(names(chi_by_q), function(nm) {
    data.table(q = as.numeric(sub("q", "", nm)), chi_by_q[[nm]])
  })
)

ggplot(q_curves, aes(x = topic, y = OpTop, color = factor(q))) +
  geom_line(linewidth = 0.7) +
  labs(x = "Topics", y = expression(OpTop[J]^{K}), color = "q",
       title = "The envelope parameter shifts the level, not the story") +
  theme_bw()
```

Applying the two selection rules to each element of the sweep — the same
one-liners as before — shows how the picks react to `q`:

```{r q-picks}
picks_by_q <- data.table(
  q = as.numeric(sub("q", "", names(chi_by_q))),
  sequential = sapply(chi_by_q, function(tab) tab[pval > alpha, topic][1]),
  min = sapply(chi_by_q, function(tab) tab[which.min(OpTop), topic])
)

knitr::kable(picks_by_q, caption = "Selected K by rule as q varies")
```

Larger `q` keeps more words per document, so the degrees of freedom grow and
the envelope covers more of each document's distribution; smaller `q` makes
the comparison coarser and cheaper. The sequential pick moves with `q` while
the global minimum stays put. The default `q = 0.95` follows the paper.

## A Note On Computational Efficiency

```{r bench-load, include = FALSE}
bench <- readRDS(system.file("extdata", "optop-benchmark-results.rds",
                             package = "OpTop"))
bres <- as.data.table(bench$results)
bres[, config := ifelse(mode == "boot_r", "R bootstrap (0.11.0)",
                        sprintf("%s, %d thread%s",
                                ifelse(mode == "stat", "statistic", "C++ bootstrap"),
                                threads, ifelse(threads > 1, "s", "")))]
```

Both computational stages of `optimal_topic()` are implemented in compiled
code and, as of v0.12.0, run in parallel. This section describes the
implementation and reports a simulation study of its speed, thread scaling,
and memory use. It concerns `optimal_topic()` only; the discrepancy indices
of the next section are vectorized R code whose cost is dominated by BLAS
operations.

**The statistic.** For each model, the fitted word probabilities of a block
of 256 documents are obtained with a single BLAS matrix product, since the
per-document algebra of Equation 8 is one row of $\gamma\,e^{\beta}$. The
remaining per-document work (a descending sort, a cumulative sum, and a
scalar Pearson reduction), which profiling identified as the dominant cost
once the algebra had been reduced to a matrix product, is parallelized over
the documents of each block with OpenMP.

**The bootstrap.** Calibration is the most expensive stage: it draws $B$
null replicates of the statistic per model, each requiring one multinomial
draw per document. Since v0.12.0 the draws are generated in compiled code
and fused with the Pearson reduction, so that no intermediate count matrices
are materialized, and the document loop is parallelized. The sampler is
chosen per document from the data alone: when the envelope holds more bins
than the document holds tokens, the common regime on large vocabularies,
the tokens are drawn directly through an alias table at $O(N_j)$ per
replicate and the untouched bins contribute a closed-form Pearson term;
otherwise the conditional-binomial method is used. Each document draws
from a private random-number stream seeded deterministically from the pair
(`seed`, document index), and all reductions are performed in a fixed order.
Two properties follow: the output is bit-identical for every value of
`n_threads`, which therefore affects wall time only, and the calibrated
p-values are reproducible under `seed`.

The study below evaluates grids of `r length(bench$meta$K_grid)` VEM models
on three synthetic corpus scales, timing the statistic pass and the
bootstrap-calibrated pass (B = `r bench$meta$n_boot`) for 1 to 4 threads,
with the pure-R bootstrap of v0.11.0 as baseline. Peak memory is the maximum
resident set size reported by the operating system for a fresh R process per
configuration. The study is reproducible via
`data-raw/benchmark-efficiency.R`; timings are machine-dependent (recorded
here: `r bench$meta$cpu`, `r bench$meta$cores` cores,
`r bench$meta$r_version`).

```{r bench-speed-table, echo = FALSE}
speed <- dcast(bres, scale + J + W ~ config, value.var = "elapsed_sec")
setorder(speed, J)
knitr::kable(speed, digits = 1,
             caption = paste("Wall time (seconds) of optimal_topic() across",
                             "the grid: statistic pass and bootstrap-calibrated",
                             "pass, by thread count"))
```

```{r bench-scaling-plot, echo = FALSE}
boot_runs <- bres[mode == "boot"]
boot_runs[, speedup := elapsed_sec[threads == 1] / elapsed_sec, by = scale]
baseline_r <- bres[mode == "boot_r", .(scale, base_r = elapsed_sec)]
boot_runs <- merge(boot_runs, baseline_r, by = "scale")
boot_runs[, vs_R := base_r / elapsed_sec]

ggplot(boot_runs, aes(x = threads, y = vs_R, color = scale)) +
  geom_line(linewidth = 0.7) +
  geom_point() +
  geom_hline(yintercept = 1, linetype = 3) +
  scale_x_continuous(breaks = sort(unique(boot_runs$threads))) +
  labs(x = "Threads", y = "Speedup over the 0.11.0 R bootstrap",
       color = "Corpus scale",
       title = "Speedup of the calibrated pass",
       subtitle = "Total speedup over the pre-0.12.0 R implementation, by thread count") +
  theme_bw()
```

```{r bench-ram, echo = FALSE}
ram <- bres[mode %in% c("boot", "boot_r")]
ram[, corpus := sprintf("J = %d, W = %d", J, W)]
ram[, corpus := factor(corpus, levels = unique(corpus[order(J)]))]
ram[, config := factor(config, levels = c("R bootstrap (0.11.0)",
                                          grep("C\\+\\+", unique(config),
                                               value = TRUE)))]

ggplot(ram, aes(x = corpus, y = max_rss_mb, fill = config)) +
  geom_col(position = "dodge") +
  labs(x = NULL, y = "Peak resident set size (MB)", fill = NULL,
       title = "Peak memory of the calibrated pass",
       subtitle = "Maximum resident set size of a fresh R process per configuration") +
  theme_bw() +
  theme(legend.position = "bottom")
```

Three regularities hold across the scales. First, the compiled bootstrap is
`r sprintf("%.1f", min(bres[mode == "boot_r", elapsed_sec] / bres[mode == "boot" & threads == 1, elapsed_sec]))`–`r sprintf("%.1f", max(bres[mode == "boot_r", elapsed_sec] / bres[mode == "boot" & threads == 1, elapsed_sec]))`×
faster than the R implementation before any threading, and the gain grows
with the vocabulary: the alias-table path costs $O(N_j)$ per replicate
where the bin-wise method costs $O(k_j)$, and wide corpora have
$k_j \gg N_j$. Second, the parallel
document loop scales almost linearly: at 4 threads the calibrated pass is
`r sprintf("%.1f", max(boot_runs[threads == 4, vs_R]))`× faster than the
0.11.0 baseline on the largest corpus, falling from
`r sprintf("%.1f", bres[scale == "large" & mode == "boot_r", elapsed_sec] / 60)` minutes
to `r sprintf("%.0f", bres[scale == "large" & mode == "boot" & threads == 4, elapsed_sec])` seconds.
The statistic pass shares the document-level parallelism and gains
`r sprintf("%.1f", bres[scale == "large" & mode == "stat" & threads == 1, elapsed_sec] / bres[scale == "large" & mode == "stat" & threads == 4, elapsed_sec])`×
at 4 threads on the same corpus; its speedup is sub-linear because the BLAS
product remains serial. Third, peak
memory is governed by the fitted models held in memory rather than by the
evaluation itself: the fused reduction yields a visible saving only at the
smallest scale, and the per-thread state amounts to one result slot per
document, so peak memory does not grow with `n_threads`. On toolchains
without OpenMP support (notably the default macOS compiler) the cores run
single-threaded; the thread-independent speedup of the compiled bootstrap
is retained, while the scaling with `n_threads` is not. The internal helper
`OpTop:::optop_openmp_available()` reports whether the installed build
supports OpenMP.

## Goodness-of-fit: the discrepancy indices

The second family of tools asks a complementary question: not *which K is
optimal* but *how much of the word-count structure a given model explains*.
The indices compare each model against the no-topics corpus baseline on a
**harmonized support**: a fixed, document-specific set of rare words
(computed across the whole grid) is collapsed so that every K is evaluated on
the same bins. They consume raw counts:

```{r indices-call, eval = FALSE}
dtm_counts <- methods::as(mydfm_sub, "CsparseMatrix")
partition <- optop_make_partition(VEM_models, dtm_counts, c = 5)
baseline <- optop_make_baseline(dtm_counts)

idx_tab <- optop_index_table(VEM_models, dtm_counts,
                             metrics = c("se", "chisq", "deviance"),
                             partition = partition, baseline = baseline,
                             macro = TRUE)
```

```{r indices-plot}
idx <- melt(idx_tab,
            id.vars = "K",
            measure.vars = c("R2_SE", "R2_chisq", "R2_dev"),
            variable.name = "metric", value.name = "R2")

ggplot(idx, aes(x = K, y = R2, color = metric)) +
  geom_line(linewidth = 0.7) +
  labs(x = "Topics", y = expression(R^2),
       title = "Discrepancy indices across the grid (document level, micro)") +
  theme_bw()
```

All three indices are regression-style R-squared analogues: 0 means the model
explains no more than the corpus-wide baseline, 1 means a perfect account of
the (harmonized) word counts. They typically rise steeply at small K and
flatten once additional topics stop buying explanatory power — the elbow
region is where the Test 1 machinery above places the optimum.

The indices also exist at the **word level**, which is useful diagnostics:
which words does the selected model capture well, and which does it miss?
For the sequential pick (K = `r word_r2$k`), the micro word-level chi-square
index is `r round(word_r2$micro, 3)`; the tails look like this:

```{r word-snapshot}
knitr::kable(
  data.frame(
    `best words` = names(word_r2$best),
    R2_best = round(unname(word_r2$best), 3),
    `worst words` = names(word_r2$worst),
    R2_worst = round(unname(word_r2$worst), 3),
    check.names = FALSE
  ),
  caption = sprintf("Word-level chi-square R2 at K = %d: the 10 best and worst captured words",
                    word_r2$k)
)
```

## A simulation with known truth

Real corpora never reveal their true K. To see the machinery work where the
answer is known, we simulate one corpus from the LDA generative process
itself: `r meta$J_sim` documents over `r meta$W_sim` words drawn from
**`r meta$true_K` topics**, using `sim_dfm()` (a thin wrapper around
`LDATS::sim_LDA_data()`), then re-fit a grid around the truth.

```{r sim-call, eval = FALSE}
rdirichlet <- function(n, alpha) {
  g <- matrix(rgamma(n * length(alpha), shape = alpha, rate = 1),
              nrow = n, byrow = TRUE)
  g / rowSums(g)
}

set.seed(20260703)
true_K <- 10L
Theta_true <- rdirichlet(100L, rep(0.3, true_K))    # documents x topics
Beta_true <- rdirichlet(true_K, rep(0.05, 1500L))   # topics x words
doc_len <- sample(500:1500, 100L, replace = TRUE)

sim_corpus <- sim_dfm(DTW = Theta_true, TWW = Beta_true,
                      doc_length = doc_len, seed = 42)

sim_models <- parallel::mclapply(
  seq(2, 20, by = 2),
  function(k) topicmodels::LDA(sim_corpus, k = k,
                               control = list(seed = 2000 + k)),
  mc.cores = ncores
)

sim_chi <- optimal_topic(sim_models,
                         quanteda::dfm_weight(sim_corpus, scheme = "prop"),
                         q = 0.95, alpha = alpha, verbose = TRUE)
```

```{r sim-plot}
ggplot(sim_chi, aes(x = topic, y = OpTop)) +
  geom_line(linewidth = 0.8, color = "royalblue") +
  geom_vline(xintercept = meta$true_K, color = "red", linetype = 1) +
  geom_vline(xintercept = sim_picks$sequential, linetype = 2) +
  geom_vline(xintercept = sim_picks$min, linetype = 3) +
  labs(x = "Topics", y = expression(OpTop[J]^{K}),
       title = "Simulated corpus: the truth is K = 10 (solid red)",
       subtitle = sprintf("sequential pick: K = %d (dashed) | global minimum: K = %d (dotted)",
                          sim_picks$sequential, sim_picks$min)) +
  theme_bw()
```

```{r sim-indices}
sim_idx <- melt(sim_idx_tab,
                id.vars = "K",
                measure.vars = c("R2_SE", "R2_chisq", "R2_dev"),
                variable.name = "metric", value.name = "R2")

ggplot(sim_idx, aes(x = K, y = R2, color = metric)) +
  geom_line(linewidth = 0.7) +
  geom_vline(xintercept = meta$true_K, color = "red", linetype = 1) +
  labs(x = "Topics", y = expression(R^2),
       title = "Discrepancy indices on the simulated corpus") +
  theme_bw()
```

The statistic bends in the neighborhood of the true K and the indices stop
improving there — the same qualitative behavior the paper's simulation study
reports (its Figure 7). As the paper notes, the identified optimum is not
always exactly `K_sim` but typically adjacent to it: here the sequential rule
stops at K = `r sim_picks$sequential` with the truth at `r meta$true_K`,
while the global minimum drifts to the grid edge as the curve keeps creeping
down past the bend. Note that this is the *uncalibrated* sequential rule
performing well: on this corpus the post-truth plateau of the standardized
statistic sits below 1, so the implicit reduced-chi-square threshold of the
saturated p-values lands close to the actual null level — the favorable case
discussed in the calibration section. Treat the tools as locating the
*region* of adequate K, with the discrepancy indices quantifying what each
candidate buys you.

## Notes

- The legacy stability helpers (`topic_stability()`, `agg_topic_stability()`,
  `agg_document_stability()`, `get_topic_models()`) are deprecated and will
  be removed in a future release; the discrepancy indices shown here are
  their replacement.
- `selection = "legacy"` reproduces the pre-0.9.9 computation exactly for
  back-compatibility and is also deprecated.
- The chi-square reference for Test 1 is a yardstick, not an exact null law;
  a calibration layer (parametric-bootstrap null) is on the roadmap — see the
  package's `AUDIT.md` for the technical background.

## References

Haldane, J. B. S. (1937). The exact value of the moments of the distribution
of chi-square. *Biometrika*, 29, 133–143.

Lewis, C. M. and Grossetti, F. (2022). A statistical approach for optimal
topic model identification. *Journal of Machine Learning Research*, 23(58),
1–20. <https://jmlr.org/papers/v23/19-297.html>
