---
title: "Working With OpTop"
subtitle: "Parametric Tests and Discrepancy Indices for Topic Models"
author: "Francesco Grossetti"
date: "July 6, 2026"
output:
  rmarkdown::html_vignette:
    toc: true
    math_method: katex
vignette: >
  %\VignetteIndexEntry{Working With OpTop}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<style>
/* the html_vignette theme styles section h4s with a left margin and a bottom
   border; the author/date lines are h4s too and would inherit both */
h4.author, h4.date { margin-left: 0; border-bottom: none; }
/* wide display equations scroll horizontally instead of clipping */
.katex-display { overflow-x: auto; overflow-y: hidden; }
</style>

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.2,
  fig.align = "center"
)
library(OpTop)
library(data.table)
library(ggplot2)

# a cli progress bar redraws itself with carriage returns, which knitr
# flattens into one very long line; the alerts are kept and the bar is
# disabled for the render
options(cli.progress_show_after = Inf)
# the captured console streams of the replay path contain the bar's
# snapshots; they are filtered so that both paths produce the same output
show_console <- function(x) {
  cat(grep("^Processing", x, invert = TRUE, value = TRUE), sep = "\n")
}

# Precomputed results shipped with the package (a few KB): tables derived from
# the full fitted grid, plus the captured console output of the optimal_topic()
# calls below.
res <- readRDS(system.file("extdata", "optop-vignette-results.rds",
                           package = "OpTop"))
meta <- res$meta
idx_tab <- res$inaugural$index_table
word_r2 <- res$inaugural$word_snapshot
ho_res <- res$inaugural$holdout
sim_chi <- res$sim$chi
sim_picks <- res$sim$picks
sim_idx_tab <- res$sim$index_table

# The fitted grid itself (~170 MB) cannot ship with the package and lives in
# data-raw/ on the maintainer's machine. When it is available, every
# optimal_topic() call below runs live; when it is not (e.g. R CMD check
# re-building the vignette from the pruned tarball), the same objects are
# assigned from the shipped results and the calls' captured console output is
# replayed, so the rendered document is the same either way.
models_path <- file.path("..", "data-raw", "VEM_models_inaugural.rds")
have_models <- file.exists(models_path)
if (have_models) {
  VEM_models <- readRDS(models_path)
  toks <- quanteda::data_corpus_inaugural |>
    quanteda::tokens(remove_punct = TRUE,
                     remove_symbols = TRUE,
                     remove_numbers = TRUE) |>
    quanteda::tokens_tolower() |>
    quanteda::tokens_remove(quanteda::stopwords())
  mydfm_sub <- quanteda::dfm_trim(quanteda::dfm(toks), min_termfreq = 5)
  weighted_dfm <- quanteda::dfm_weight(mydfm_sub, scheme = "prop")
} else {
  chi_seq <- res$inaugural$chi_by_q[["q0.95"]]
  chi_min <- data.table::copy(chi_seq)
  chi_by_q <- res$inaugural$chi_by_q
  chi_cal <- res$inaugural$chi_cal
  chi_cal_mm <- res$inaugural$chi_cal_mm
}
```

OpTop implements the parametric tests of Lewis and Grossetti (2022, JMLR
23(58)) for identifying the optimal number of topics of a Latent Dirichlet
Allocation (LDA) model, together with a family of regression-style
goodness-of-fit (GoF) indices. This vignette illustrates both on the corpus
of the paper's case study and concludes with a simulation study in which the
true number of topics is known.

The only step performed offline is the estimation of the LDA models:
`r length(meta$K_grid)` fits on the full corpus and
`r length(ho_res$split$K_grid)` fits on the training split of the held-out
analysis. Everything downstream, from every `optimal_topic()` call to each
table, figure, and number quoted in the text, is computed from those fits.

## Setting Up: Corpus, Grid, Models

We use the U.S. Presidential Inaugural Address corpus distributed with
**quanteda**, the same corpus analyzed in the paper's case study, with a
standard preprocessing pipeline and a vocabulary trim. The grid spans
K = `r min(meta$K_grid)` to `r max(meta$K_grid)` in steps of 2.

```{r fitting, eval = FALSE}
library(OpTop)

ncores <- 6

# Tokenize
toks <- quanteda::data_corpus_inaugural |>
  quanteda::tokens(remove_punct = TRUE,
                   remove_symbols = TRUE,
                   remove_numbers = TRUE) |>
  quanteda::tokens_tolower() |>
  quanteda::tokens_remove(quanteda::stopwords())

# Create DFM (with raw counts as well as proportions)
mydfm        <- quanteda::dfm(toks)
mydfm_sub    <- quanteda::dfm_trim(mydfm, min_termfreq = 5)
weighted_dfm <- quanteda::dfm_weight(mydfm_sub, scheme = "prop")

K_grid <- seq(2, 200, by = 2)

# Estimate VEM models in parallel
VEM_models <- parallel::mclapply(
  K_grid,
  function(k) {
    message("Estimating LDA with K = ", k)
    topicmodels::LDA(x = mydfm_sub, k = k)
  },
  mc.cores = ncores
)
```

After trimming, the corpus has `r meta$ndoc` documents and
`r meta$nfeat_trimmed` features (down from `r meta$nfeat_full`). The VEM fits
above are unseeded, so re-running the pipeline may alter individual
statistics slightly; the numbers quoted in this vignette come from one such
run.

## The Test 1 Statistic

The test evaluates overall model adequacy: how well a $K$-topic model
explains the corpus. It relies on the multinomial structure that LDA itself
assumes. A fitted $K$-topic model describes document $j$ by a vector of
document-topic weights $\theta_j^K$ and each topic by a distribution over
the $P$ vocabulary words, collected in the topic-word matrix $\phi^K$. The
model's fitted probability of word $p$ in document $j$ is

$$
I_{jp}^{K} = \sum_{k=1}^{K} \theta_{jk}^{K}\,\phi_{kp}^{K},
\tag{1}
$$

which is Equation 5 in the paper. Equation 4 of the paper represents the
observed word-proportion vector $d_j$ as this fitted distribution plus
multinomial sampling noise. If the $K$-topic model fully characterizes the
corpus, one would be unable to reject the hypothesis that the observed and
estimated word distributions are statistically indistinct: a classical
goodness-of-fit situation.

Because a large number of words carry near-zero fitted probability, the
comparison is made on a per-document **envelope**. The fitted probabilities
$I_{jp}^K$ are sorted in decreasing order and the smallest leading set whose
cumulative mass exceeds $q$ is retained; these are the $P_j$ relatively
important words of document $j$. The remaining, relatively unimportant words
are collapsed into a single bin (the paper's Equations 6 and 7). With the
words of document $j$ indexed in decreasing order of $I_{jp}^K$, the fitted
and observed masses of that bin are

$$
I^{K}_{j,\min} = \sum_{p = P_j + 1}^{P} I^{K}_{jp} \;<\; 1 - q,
\tag{2a}
$$

$$
D_{j,\min} = \sum_{p = P_j + 1}^{P} D_{jp}.
\tag{2b}
$$

Collapsing the relatively unimportant words in this manner preserves the
underlying multinomial assumption. Note that the basis for the cutoff is the
fitted model rather than the observed documents; the paper's numerical
setting $I^K = 0.05$ corresponds to `q = 0.95`. Test 1 (Equation 8 of the
paper) then aggregates the Pearson discrepancies over documents:

$$
\mathrm{OpTop}_J^K = \sum_{j=1}^{J} 
\left[\, (P_j + 1) \left(
\sum_{p=1}^{P_j} \frac{\bigl(D_{jp} - I_{jp}^K\bigr)^2}{I_{jp}^K}
+\; \frac{\bigl(D_{j,\min} - I_{j,\min}^K\bigr)^2}{I_{j,\min}^K}
\right) \right] \sim \chi^2_{P_J},
\tag{3}
$$

where $\mathrm{OpTop}_J^K$ is distributed chi-square with
$P_J = \sum_{j=1}^{J} P_j$ degrees of freedom. Misspecification, that is a
$K$ too coarse to reproduce the word distributions, can only inflate the
statistic, so evidence against adequacy is found in the upper tail. The
package reports the *standardized* statistic $\mathrm{OpTop}_J^K / P_J$
(the version plotted in the paper's Figure 2, which fluctuates around 1
under an adequate model) in the `OpTop` column, the degrees of freedom $P_J$
in `df`, and the upper-tail p-value in `pval`.

## Running the Test

`optimal_topic()` consumes *word proportions* (`weighted_dfm` above),
while the discrepancy indices of the second part consume the raw *counts*;
both objects are therefore required. With `verbose = TRUE`, the function
reports its setup, its progress across the grid, the selected model together
with the rule that chose it, and the elapsed time:

```{r optimal-call-seq, eval = have_models}
# Use the sequential approach
chi_seq <- optimal_topic(topic_models = VEM_models,
                         weighted_dfm = weighted_dfm,
                         selection = "sequential",
                         q = 0.95,
                         alpha = 0.01,
                         do_plot = FALSE)
```

```{r optimal-call-seq-replay, eval = !have_models, echo = FALSE}
show_console(res$inaugural$console_seq)
```

```{r optimal-call-min, eval = have_models}
# Use the global minimum approach
chi_min <- optimal_topic(topic_models = VEM_models,
                         weighted_dfm = weighted_dfm,
                         selection = "min",
                         q = 0.95,
                         alpha = 0.01,
                         do_plot = FALSE)
```

```{r optimal-call-min-replay, eval = !have_models, echo = FALSE}
show_console(res$inaugural$console_min)
```

## Reading the Result

`optimal_topic()` returns a `data.table` with one row per model:

```{r table-head}
head(chi_seq)
```

- `OpTop` is the *standardized* Test 1 statistic, the raw chi-square of
  Equation (3) divided by its degrees of freedom, and the quantity plotted
  in the paper's Figure 2. Under an adequate model it fluctuates around 1.
- `df` is $P_J = \sum_j P_j$, the total number of relatively important words
  across documents.
- `pval` is the upper-tail p-value of the raw statistic. Small values
  indicate that the observed discrepancy is larger than sampling noise can
  explain, so the $K$-topic model is rejected.

The `selection` rule affects which K is *reported* as optimal (and the plot
guides), never the table itself. The two calls above return identical
tables:

```{r rule-independence}
all.equal(chi_seq, chi_min)
```

### The Two Selection Rules

Both rules are direct computations on the returned table and can be stated
explicitly:

```{r selection-rules}
alpha <- 0.01

# "sequential": the smallest K the test fails to reject
seq_pick <- chi_seq[pval > alpha, topic][1]

# "min": the K with the minimum standardized statistic
min_pick <- chi_seq[which.min(OpTop), topic]

c(sequential = seq_pick, min = min_pick)
```

**`selection = "sequential"`** implements the classical sequential scan for
model order: $K$ is increased until the first model the test *fails to
reject* (`pval > alpha`). The rationale is parsimony: the rule selects the
smallest $K$ that the data cannot distinguish from adequate. On this corpus
the sequential rule selects **K = `r seq_pick`**. If every model were
rejected, `optimal_topic()` would fall back to the global minimum with a
warning.

**`selection = "min"`** reproduces the published case study: the optimum is
identified as the $K$-topic model with the minimum standardized statistic
across the grid. Here the minimum is at **K = `r min_pick`**, close to the
92 topics the paper identifies on the untrimmed vocabulary of the same
corpus and, as the sweep below shows, insensitive to the choice of `q`. One
practical remark is in order: the minimum rule requires a grid that extends
beyond the flattening of the curve. On a grid truncated at K = 100 the
minimum sits at the boundary, and only the full scan up to K = 200 allows it
to stabilize. The behavior has a structural explanation. The `df` column
decreases steadily with $K$ because the envelopes shrink as the fitted
distributions concentrate, and since the null level of the standardized
statistic scales with the envelope size, part of the curve's long decline
reflects envelope shrinkage rather than genuine improvement in fit. The
calibrated p-values of the next section absorb this effect automatically;
the raw curve does not. The sequential rule, by construction, stops at the
first adequate $K$ and is unaffected by the extent of the grid.

The p-value column explains when and why the two rules diverge:

```{r pval-summary}
summary(chi_seq$pval)
```

With `df` of the order of
`r format(round(median(chi_seq$df)), big.mark = ",")`, the chi-square
reference distribution is extremely concentrated, so p-values saturate:
models whose standardized statistic is materially above 1 receive p-values
of essentially 0, and models at or below 1 receive p-values of essentially
1. When every model is rejected, the sequential rule falls back to the
global minimum with a warning; when several models are accepted, it stops at
the smallest, which may precede the minimum by a wide margin. This is a
documented calibration caveat of the chi-square reference (see
`?optimal_topic`): in saturated regimes the information is carried by the
*shape* of the `OpTop` curve and by its minimum.

Because the return value is a table, custom figures follow directly:

```{r optop-curve}
ggplot(chi_seq, aes(x = topic, y = OpTop)) +
  geom_line(linewidth = 0.8, color = "royalblue") +
  geom_vline(xintercept = seq_pick, linetype = 2) +
  geom_vline(xintercept = min_pick, linetype = 3) +
  annotate("point", x = min_pick, y = min(chi_seq$OpTop),
           color = "red", shape = 4, size = 4) +
  labs(x = "Topics", y = expression(OpTop[J]^{K}),
       title = "Standardized Test 1 statistic across the grid",
       subtitle = sprintf("sequential pick: K = %d (dashed) | global minimum: K = %d (dotted)",
                          seq_pick, min_pick)) +
  theme_bw()
```

## Calibrated p-Values

The saturation above has a precise cause: the $\chi^2_{P_J}$ reference is an
asymptotic guide rather than the exact null law of the statistic. Write
$k_j = P_j + 1$ for the number of envelope bins of document $j$ and $N_j$
for its length in tokens. The classical Pearson statistic on *counts*,
$X^2_j = N_j \sum_b (d_{jb} - p_{jb})^2 / p_{jb}$, is the object with the
textbook $\chi^2_{k_j - 1}$ asymptotics; the document term of Equation (3)
is instead

$$
S_j = (P_j + 1) \sum_b \frac{(d_{jb} - p_{jb})^2}{p_{jb}}
= \frac{k_j}{N_j}\, X^2_j ,
\tag{4}
$$

which differs from the calibrated object by the factor $k_j / N_j$. In
addition, the fitted probabilities are estimated from the same data they are
tested against. Calibration therefore replaces the $\chi^2$ reference with
the distribution of $T = \sum_j S_j$ under the **conditional fitted-model
null**

$$
c_j \sim \mathrm{Multinomial}\bigl(N_j,\; I_j^K\bigr),
\qquad j = 1, \dots, J,
\tag{5}
$$

with $\hat\theta$ and $\hat\phi$ held fixed at their estimates. Two
properties make this exact and fast: the envelope depends only on the model,
never on the data, and a multinomial collapsed over bins is multinomial on
the collapsed probabilities. The null replicates can therefore be drawn
*directly on the $k_j$ bins*
$p_j = (I^K_{j1}, \dots, I^K_{jP_j}, I^K_{j,\min})$, which is equivalent to
simulating entire documents over the vocabulary at a small fraction of the
cost. The only new ingredient the user supplies is the vector of document
lengths $N_j$:

```{r optimal-call-cal, eval = have_models}
doc_lens <- quanteda::ntoken(mydfm_sub)

chi_cal <- optimal_topic(topic_models = VEM_models,
                         weighted_dfm = weighted_dfm,
                         q = 0.95,
                         alpha = 0.01,
                         calibrate = "bootstrap",
                         n_boot = 200,
                         doc_lengths = doc_lens,
                         seed = 20260703,
                         do_plot = FALSE)
```

```{r optimal-call-cal-replay, eval = !have_models, echo = FALSE}
show_console(res$inaugural$console_cal)
```

With `calibrate = "bootstrap"`, $B$ = `n_boot` null replicates
$T^{*}_1, \dots, T^{*}_B$ are drawn as above and the reported p-value is the
empirical upper tail with the standard finite-sample correction,

$$
\hat{p} = \frac{1 + \#\{\, b : T^{*}_b \ge T \,\}}{B + 1},
\tag{6}
$$

so `alpha` is a genuine Type-I error rate with respect to the conditional
null, at resolution $1/(B+1)$. The bootstrap is the most expensive stage of
the pipeline; its cost, and the effect of the `n_threads` argument, are
quantified in the section [A Note On Computational Efficiency]. The table
now carries two p-value columns: `pval` (the calibrated value used by the
selection rules) and `pval_chisq` (the asymptotic value, kept for
comparison):

```{r cal-table}
head(chi_cal)
```

The bootstrap is the reference method; `calibrate = "moment"` targets the
same null in closed form. Haldane (1937) gives the exact multinomial moments
of the count-based Pearson statistic over $k_j$ bins:

$$
\mathrm{E}\bigl[X^2_j\bigr] = k_j - 1,
\tag{7a}
$$

$$
\begin{split}
\mathrm{Var}\bigl[X^2_j\bigr] &= 2(k_j - 1)
\\
&\quad + \frac{\sum_b 1/p_{jb} - k_j^2 - 2k_j + 2}{N_j}.
\end{split}
\tag{7b}
$$

Since $S_j = (k_j/N_j)\,X^2_j$ and documents are independent, the null mean
and variance of $T$ follow by scaling and summing,
$\mu = \sum_j (k_j/N_j)(k_j - 1)$ and
$\sigma^2 = \sum_j (k_j/N_j)^2\,\mathrm{Var}[X^2_j]$, and the reference is
the Satterthwaite scaled chi-square matching them:

$$
\begin{gathered}
T \overset{\text{approx}}{\sim} a\,\chi^2_\nu,
\qquad a = \frac{\sigma^2}{2\mu},
\qquad \nu = \frac{2\mu^2}{\sigma^2},
\\
\hat{p} = \Pr\!\left(\chi^2_\nu \ge T/a\right).
\end{gathered}
\tag{8}
$$

The method requires no simulation and no seed; it is a fast approximation
that corrects the location and the scale of the reference, although not its
higher moments:

```{r optimal-call-mm, eval = have_models}
chi_cal_mm <- optimal_topic(topic_models = VEM_models,
                            weighted_dfm = weighted_dfm,
                            q = 0.95,
                            alpha = 0.01,
                            calibrate = "moment",
                            doc_lengths = doc_lens,
                            do_plot = FALSE)
```

```{r optimal-call-mm-replay, eval = !have_models, echo = FALSE}
show_console(res$inaugural$console_cal_mm)
```

The following figure compares the three reference distributions:

```{r cal-plot}
pv <- melt(
  data.table(topic = chi_cal$topic,
             `chi-square` = chi_cal$pval_chisq,
             bootstrap = chi_cal$pval,
             moment = chi_cal_mm$pval),
  id.vars = "topic", variable.name = "reference", value.name = "pvalue"
)

ggplot(pv, aes(x = topic, y = pvalue, color = reference)) +
  geom_line(linewidth = 0.7) +
  geom_hline(yintercept = alpha, linetype = 3) +
  labs(x = "Topics", y = "p-value",
       title = "P-values under the three reference distributions",
       subtitle = sprintf("dotted line: alpha = %.2f", alpha)) +
  theme_bw()
```

Because the p-values themselves do not depend on `alpha`, the conventional
thresholds can be scanned directly on the tables already computed:

```{r cal-alpha-sweep}
alphas <- c(0.01, 0.05, 0.10)

picks_alpha <- data.table(
  alpha = alphas,
  bootstrap = sapply(alphas, function(a) chi_cal[pval > a, topic][1]),
  moment = sapply(alphas, function(a) chi_cal_mm[pval > a, topic][1]),
  `chi-square` = sapply(alphas, function(a) chi_seq[pval > a, topic][1])
)

knitr::kable(picks_alpha,
             caption = "Sequential pick by reference distribution as alpha varies")
```

Interpretation follows from the null the calibration targets. The calibrated
sequential pick answers a precise question: *what is the smallest $K$ whose
residual discrepancy could plausibly be produced by multinomial sampling
noise alone, if the fitted model were the truth?* On this corpus this
happens already at K = `r chi_cal[pval > alpha, topic][1]`. The fit does
continue to improve beyond that point, since the `OpTop` curve decreases
until its minimum at K = `r min_pick`, but from there onward the
improvements are smaller than what sampling noise can account for at the
chosen `alpha`.

The uncalibrated scan is best understood as a *reduced-chi-square
heuristic*. With `df` this large the asymptotic p-values saturate at 0 or 1,
so in practice the rule stops at the first $K$ whose standardized
discrepancy falls below 1. This is a reasonable descriptive criterion, but
its implicit threshold is not tied to the actual null level of the
statistic. On this corpus the bootstrap places that null level closer to 2
than to 1, so the heuristic demands more than noise-level adequacy and stops
later, at K = `r seq_pick`. The selection is also unstable: on this grid the
next model, K = `r seq_pick + 2`, is rejected again, and the sweep of `q`
below moves the pick from 20 to 38. When the implicit threshold happens to
lie close to the true null level, the same heuristic performs well; the
simulation study at the end of this vignette, where the ground truth is
known, is such a case. What calibration adds is therefore not a different
answer but a guarantee. Raising `alpha` demands stronger evidence of
adequacy, so the pick moves weakly upward, as the table shows, and `alpha`
becomes a genuine Type-I error rate with respect to the fitted-model null,
at bootstrap resolution `1 / (n_boot + 1)`. In summary, the calibrated
sequential rule identifies the most parsimonious *statistically adequate*
$K$, while the global-minimum rule and the discrepancy indices below
quantify *maximal descriptive fit*. The two questions are distinct.

One caveat applies to both calibrations: they hold the estimated `theta` and
`phi` fixed, so they do not account for estimation noise in the LDA
parameters themselves (see `?optimal_topic`).

## The Role of `q`

`q` controls the per-document envelope introduced in Equations (2a) and (2b): the
fitted word probabilities are sorted in decreasing order and the smallest
leading set whose cumulative mass exceeds `q` is retained, while the
relatively unimportant words are collapsed into a single bin whose mass
remains strictly below `1 - q`. In the paper's notation `q = 1 - I^K`, and
its numerical setting `I^K = 0.05` corresponds to the default `q = 0.95`.

Re-running the test across a sweep of `q` values is a direct sensitivity
check:

```{r q-sweep-run, eval = have_models}
q_grid <- c(0.80, 0.90, 0.95, 0.99)

chi_by_q <- lapply(q_grid, function(q) {
  optimal_topic(VEM_models, weighted_dfm, q = q, alpha = alpha,
                do_plot = FALSE)
})
names(chi_by_q) <- sprintf("q%.2f", q_grid)
```

```{r q-sweep-plot}
q_curves <- rbindlist(
  lapply(names(chi_by_q), function(nm) {
    data.table(q = as.numeric(sub("q", "", nm)), chi_by_q[[nm]])
  })
)

ggplot(q_curves, aes(x = topic, y = OpTop, color = factor(q))) +
  geom_line(linewidth = 0.7) +
  labs(x = "Topics", y = expression(OpTop[J]^{K}), color = "q",
       title = "Standardized Test 1 statistic across the q sweep") +
  theme_bw()
```

Applying the two selection rules to each element of the sweep shows how the
selected $K$ responds to `q`:

```{r q-picks}
picks_by_q <- data.table(
  q = as.numeric(sub("q", "", names(chi_by_q))),
  sequential = sapply(chi_by_q, function(tab) tab[pval > alpha, topic][1]),
  min = sapply(chi_by_q, function(tab) tab[which.min(OpTop), topic])
)

knitr::kable(picks_by_q, caption = "Selected K by rule as q varies")
```

Larger `q` retains more words per document, so the degrees of freedom grow
and the envelope covers more of each document's distribution, while smaller
`q` makes the comparison coarser and computationally cheaper. The sequential
pick moves with `q`, whereas the global minimum is unaffected. The default
`q = 0.95` follows the paper.

## Goodness-of-Fit: The Discrepancy Indices

The second family of tools asks a complementary question: not *which K is
optimal* but *how much of the word-count structure a given model explains*.
Following Lewis and Grossetti (2026), each index measures the proportional
reduction in discrepancy achieved by a $K$-topic model relative to the
no-topics baseline in which every document follows the corpus word
distribution. The comparison is carried out on a **harmonized support**: a
fixed, document-specific set of rare words, computed as a union over the
whole model grid *and* the baseline, is collapsed so that every $K$ is
evaluated on the same bins. Because the harmonized set depends on the grid
and on the threshold constant `c`, indices are comparable across studies
only under a common pair; the default `c = 1` follows the paper's
recommendation for deviance-primary work. The indices consume raw counts:

```{r indices-call, eval = FALSE}
dtm_counts <- methods::as(mydfm_sub, "CsparseMatrix")
partition <- optop_make_partition(VEM_models, dtm_counts, c = 1)
baseline <- optop_make_baseline(dtm_counts)

idx_tab <- optop_index_table(VEM_models, dtm_counts,
                             metrics = c("se", "chisq", "deviance"),
                             partition = partition, baseline = baseline,
                             macro = TRUE)
```

```{r indices-plot}
idx <- melt(idx_tab,
            id.vars = "K",
            measure.vars = c("R2_SE", "R2_chisq", "R2_dev"),
            variable.name = "metric", value.name = "R2")

ggplot(idx, aes(x = K, y = R2, color = metric)) +
  geom_line(linewidth = 0.7) +
  labs(x = "Topics", y = expression(R^2),
       title = "Discrepancy indices across the grid (document level, micro)") +
  theme_bw()
```

All three indices are regression-style R-squared analogues: 0 indicates that
the model explains no more than the corpus-wide baseline, and 1 indicates a
perfect account of the harmonized word counts. The deviance index is
bounded above by one, while the Pearson and squared-error indices are not
confined to $[0, 1]$; a negative value is informative, since it indicates
that the fitted model performs worse than the baseline under that
discrepancy. The indices typically rise steeply at small $K$ and flatten
once additional topics no longer provide significant incremental
explanatory power; the region where the curves flatten corresponds to the
neighborhood in which Test 1 locates the optimum. These full-corpus values
are descriptive summaries of in-sample fit and should not be the sole basis
for choosing $K$.

The indices are also available at the **word level**, computed on the
unbinned vocabulary, which serves as a diagnostic: it identifies the words
the selected model captures well and those it misses. For the sequential
pick (K = `r word_r2$k`), the micro word-level chi-square index is
`r round(word_r2$micro, 3)`; the ten highest- and lowest-index words are
reported below:

```{r word-snapshot}
knitr::kable(
  data.frame(
    `best words` = names(word_r2$best),
    R2_best = round(unname(word_r2$best), 3),
    `worst words` = names(word_r2$worst),
    R2_worst = round(unname(word_r2$worst), 3),
    check.names = FALSE
  ),
  caption = sprintf("Word-level chi-square R2 at K = %d: the ten best and worst captured words",
                    word_r2$k)
)
```

## Held-Out Validation and Topic-Number Choice

The indices above are computed on the same documents used to estimate the
models, so they describe in-sample fit and are optimistic by construction:
every additional topic gives the model more freedom to adapt to the
estimation sample, whether or not the gain reflects structure that
generalizes. Formal inference on fit, and a topic-number choice that can be
defended statistically, require documents the models have never seen
(Lewis and Grossetti, 2026).

We therefore hold out `r ho_res$split$n_eval` of the `r meta$ndoc` speeches,
chosen at random under a fixed seed, and refit the grid on the remaining
`r ho_res$split$n_train` alone, for K = `r min(ho_res$split$K_grid)` to
`r max(ho_res$split$K_grid)` in steps of 2. The vocabulary is fixed by the
training split; the `r ho_res$split$n_oov` word types that occur only in
held-out speeches (about `r round(100 * ho_res$split$oov_share, 1)`% of the
evaluation tokens) are collapsed into a synthetic out-of-vocabulary bin that
every model treats as rare. The corpus baseline is likewise estimated on the
training split, so nothing in the evaluation documents influences the fitted
models or the targets they are compared against.

```{r holdout-call, eval = FALSE}
set.seed(20260709)
n_docs <- quanteda::ndoc(mydfm_sub)
eval_idx <- sort(sample(n_docs, round(0.25 * n_docs)))
dfm_eval <- mydfm_sub[eval_idx, ]
dfm_train <- quanteda::dfm_trim(mydfm_sub[-eval_idx, ], min_termfreq = 1)

VEM_train <- lapply(seq(2, 60, by = 2), function(k) {
  topicmodels::LDA(x = dfm_train, k = k, control = list(seed = 500 + k))
})

base_train <- optop_make_baseline(dfm_train)
ho <- optop_index_holdout(VEM_train, dfm_eval, base_train)
```

`optop_index_holdout()` gives each evaluation document its topic weights
under the trained topics held fixed, the fold-in step, reconstructs its word
distribution, and scores the reconstruction with the same discrepancy
indices as above, on a rare-word partition built from the training baseline.
Because the evaluation documents are independent of the fitted models, the
per-document indices support genuine confidence statements. The figure
reports the Macro deviance index, the unweighted average of the per-document
indices, with its 95% confidence band across the grid:

```{r holdout-plot}
ho_dev <- ho_res$summary[ho_res$summary$metric == "deviance", ]

ggplot(ho_dev, aes(x = K, y = macro)) +
  geom_ribbon(aes(ymin = ci_lo, ymax = ci_hi), fill = "grey80") +
  geom_line(linewidth = 0.7) +
  labs(x = "Topics", y = expression(R^2),
       title = "Held-out Macro deviance index with 95% confidence band") +
  theme_bw()
```

The band quantifies the uncertainty in the average held-out fit across
evaluation documents, conditional on the trained models. The summary table
also reports the Micro index, which weights each document by its baseline
discrepancy, and the Micro-Macro gap with its standard error; a gap far from
zero indicates that fit quality varies systematically with document weight.

The curve flattens quickly, and reading a single best K off it would ignore
the sampling noise. `optop_gain_table()` makes the choice formal: for each
adjacent pair of topic counts it computes the paired per-document
improvement in the held-out index, its one-sided upper confidence bound, and
a test of the hypothesis that the improvement is zero.

```{r gain-call, eval = FALSE}
gt <- optop_gain_table(ho, metric = "deviance", epsilon = 0.01)
```

```{r gain-table}
show_upto <- if (identical(ho_res$selection_status, "selected")) {
  selected_gain <- match(ho_res$k_hat, ho_res$gains$K)
  min(nrow(ho_res$gains), selected_gain + 2L)
} else {
  min(nrow(ho_res$gains), 6L)
}
knitr::kable(
  ho_res$gains[seq_len(show_upto), ],
  digits = 4,
  caption = sprintf("Paired adjacent gains in the held-out deviance index (first %d of %d rows)",
                    show_upto, nrow(ho_res$gains))
)
```

Each row compares K with its successor on the same evaluation documents:
`gain` is the average paired improvement, `upper` its one-sided
`r sprintf("%d", round(100 * (1 - ho_res$alpha_gain)))`% upper confidence
bound, and `pval` tests whether the improvement exceeds zero. The selection
rule fixes a tolerance epsilon, the improvement judged too small to matter,
before looking at the results, and picks the smallest K at which even the
optimistic end of the next gain falls below it.

```{r gain-decision, echo = FALSE, results = "asis"}
if (identical(ho_res$selection_status, "selected")) {
  cat(sprintf(
    "With epsilon = %s, the rule selects **K = %d**: the step out of %d is the first whose upper bound falls below the tolerance, so no gain of practical size is compatible with the data at that step.\n\n",
    ho_res$epsilon, ho_res$k_hat, ho_res$k_hat
  ))
} else {
  cat(sprintf(
    "With epsilon = %s, no step in this grid has an upper bound below the tolerance. The moment diagnostic below therefore uses K = %d, the grid point with the largest held-out Macro index, only as an illustration and not as an adequacy selection.\n\n",
    ho_res$epsilon, ho_res$k_diagnostic
  ))
}
```

A scalar index says how much error remains, not where. The moment tests of
Lewis and Grossetti (2026) complement it by grouping the vocabulary into
labeled word sets built from the training sample, the instruments, and
testing whether the model consistently over- or under-predicts any group on
the evaluation documents. Here we stratify the vocabulary by training
frequency and compare each stratum against the highest-frequency reference:

```{r moment-call, eval = FALSE}
k_diagnostic <- gt$k_hat
if (is.na(k_diagnostic)) {
  dev <- ho$summary[ho$summary$metric == "deviance", ]
  k_diagnostic <- dev$K[which.max(dev$macro)]
}
m_star <- VEM_train[[which(seq(2, 60, by = 2) == k_diagnostic)]]
mt <- optop_moment_test(list(m_star), dfm_eval, dfm_train,
                        type = "strata", bins = 5)
```

```{r moment-table}
knitr::kable(
  ho_res$moment_marginal,
  digits = c(0, 5, 5, 2, 4, 4),
  format.args = list(scientific = FALSE),
  caption = sprintf("Marginal frequency-strata moments at K = %d",
                    ho_res$k_diagnostic)
)
```

The joint Wald statistic at K = `r ho_res$k_diagnostic` is
`r round(ho_res$moment_summary$wald, 2)` on `r ho_res$moment_summary$df`
degrees of freedom
(p `r if (ho_res$moment_summary$pval < 1e-15) "< 1e-15" else sprintf("= %.3g", ho_res$moment_summary$pval)`);
with `r ho_res$split$n_eval` evaluation documents the chi-square reference
is an asymptotic yardstick, and the marginal t statistics are the more
stable guide. In the marginal table, `estimate` is an effect size in
probability-mass units: a positive value means the held-out documents place
more mass on that frequency stratum than the model predicts, relative to
the reference stratum, and a negative value the reverse. Here the largest
imbalance sits in the lowest-frequency stratum, with an estimate of about
`r sprintf("%.5f", ho_res$moment_marginal$estimate[which.max(abs(ho_res$moment_marginal$t))])`
in probability mass: statistically unambiguous, yet small in practical
terms. Reading the estimates alongside the p-values in this way separates
detectable imbalances from practically negligible ones.

Three caveats close the section. First, all confidence statements are
conditional on the trained models: they quantify variability across
evaluation documents, not across refits of the grid. Second, selecting K on
this same evaluation sample, or choosing a diagnostic K from it when the
adequacy rule does not select one, makes the reported diagnostic descriptive
and subject to post-selection bias. A study that needs both a data-driven
choice and clean inference at that choice should select on one evaluation
sample and report on another, or rotate the roles of the splits. Third, a
moment test that fails to reject does not certify the model class; it states
that no systematic residual imbalance is detectable along the chosen
instruments at the available sample size.

## A Simulation with Known Truth

In empirical corpora the true number of topics is unknown. Following the
paper, we therefore validate the procedure through a simulation study in
which the ground truth is known: one corpus is simulated from the LDA
generative process itself, with `r meta$J_sim` documents over
`r meta$W_sim` words drawn from **`r meta$true_K` topics**, using
`sim_dfm()` (a thin wrapper around `LDATS::sim_LDA_data()`), and a grid of
models is re-estimated around the truth.

```{r sim-call, eval = FALSE}
rdirichlet <- function(n, alpha) {
  g <- matrix(rgamma(n * length(alpha), shape = alpha, rate = 1),
              nrow = n, byrow = TRUE)
  g / rowSums(g)
}

set.seed(20260703)
true_K <- 10L
Theta_true <- rdirichlet(100L, rep(0.3, true_K))    # documents x topics
Beta_true <- rdirichlet(true_K, rep(0.05, 1500L))   # topics x words
doc_len <- sample(500:1500, 100L, replace = TRUE)

sim_corpus <- sim_dfm(DTW = Theta_true, TWW = Beta_true,
                      doc_length = doc_len, seed = 42)

sim_models <- parallel::mclapply(
  seq(2, 20, by = 2),
  function(k) topicmodels::LDA(sim_corpus, k = k,
                               control = list(seed = 2000 + k)),
  mc.cores = ncores
)

sim_chi <- optimal_topic(sim_models,
                         quanteda::dfm_weight(sim_corpus, scheme = "prop"),
                         q = 0.95, alpha = alpha, verbose = TRUE)
```

```{r sim-plot}
ggplot(sim_chi, aes(x = topic, y = OpTop)) +
  geom_line(linewidth = 0.8, color = "royalblue") +
  geom_vline(xintercept = meta$true_K, color = "red", linetype = 1) +
  geom_vline(xintercept = sim_picks$sequential, linetype = 2) +
  geom_vline(xintercept = sim_picks$min, linetype = 3) +
  labs(x = "Topics", y = expression(OpTop[J]^{K}),
       title = "Simulated corpus with ground truth K = 10 (solid red)",
       subtitle = sprintf("sequential pick: K = %d (dashed) | global minimum: K = %d (dotted)",
                          sim_picks$sequential, sim_picks$min)) +
  theme_bw()
```

```{r sim-indices}
sim_idx <- melt(sim_idx_tab,
                id.vars = "K",
                measure.vars = c("R2_SE", "R2_chisq", "R2_dev"),
                variable.name = "metric", value.name = "R2")

ggplot(sim_idx, aes(x = K, y = R2, color = metric)) +
  geom_line(linewidth = 0.7) +
  geom_vline(xintercept = meta$true_K, color = "red", linetype = 1) +
  labs(x = "Topics", y = expression(R^2),
       title = "Discrepancy indices on the simulated corpus") +
  theme_bw()
```

The statistic bends in the neighborhood of the true $K$ and the indices stop
improving there, the same qualitative behavior reported in the paper's
simulation study (its Figure 7). As the paper acknowledges, the identified
optimum is not always exactly $K_{sim}$ but is typically adjacent or in
close proximity to it: here the sequential rule stops at
K = `r sim_picks$sequential` with the truth at `r meta$true_K`, while the
global minimum moves toward the grid boundary as the curve continues to
decrease past the bend. Note that this is the *uncalibrated* sequential rule
performing well. On this corpus the post-truth plateau of the standardized
statistic lies below 1, so the implicit reduced-chi-square threshold of the
saturated p-values falls close to the actual null level, which is the
favorable case discussed in the calibration section. The tools should be
read as locating the *region* of adequate $K$, with the discrepancy indices
quantifying the incremental explanatory power of each candidate.

## A Note On Computational Efficiency

```{r bench-load, include = FALSE}
bench <- readRDS(system.file("extdata", "optop-benchmark-results.rds",
                             package = "OpTop"))
bres <- as.data.table(bench$results)
scale_labels <- c(small = "small", medium1 = "medium-1",
                  medium2 = "medium-2", large = "large")
bres[, scale := factor(scale_labels[scale], levels = scale_labels)]
thread_label <- sprintf("%d thread%s", bres$threads,
                        ifelse(bres$threads > 1, "s", ""))
bres[, config := fcase(
  mode == "boot_r", "R bootstrap (0.11.0)",
  mode == "indices_r", "R implementation (pre-engine)",
  mode == "stat", paste0("statistic, ", thread_label),
  mode == "boot", paste0("C++ bootstrap, ", thread_label),
  mode == "indices", paste0("engine, ", thread_label)
)]
```

Both computational stages of `optimal_topic()` are implemented in compiled
code and, as of v0.12.0, run in parallel. The same design extends to the
goodness-of-fit tools presented above, whose partition and index
computations are evaluated by a fused compiled engine. This section
describes the implementations and reports a simulation study of their
speed, thread scaling, and memory use.

**The statistic.** For each model, the fitted word probabilities of a block
of 256 documents are obtained with a single BLAS matrix product, since the
per-document algebra of Equation (3) reduces to one row of the matrix
product $\theta^K \phi^K$. The remaining per-document work (a descending
sort, a cumulative sum, and a scalar Pearson reduction), which profiling
identified as the dominant cost once the algebra had been reduced to a
matrix product, is parallelized over the documents of each block with
OpenMP.

**The bootstrap.** Calibration is the most expensive stage: it draws $B$
null replicates of the statistic per model, each requiring one multinomial
draw per document. Since v0.12.0 the draws are generated in compiled code
and fused with the Pearson reduction, so that no intermediate count matrices
are materialized, and the document loop is parallelized. The sampler is
chosen per document from the data alone: when the envelope holds more bins
than the document holds tokens, the common regime on large vocabularies,
the tokens are drawn directly through an alias table at $O(N_j)$ per
replicate and the untouched bins contribute a closed-form Pearson term;
otherwise the conditional-binomial method is used. Each document draws
from a private random-number stream seeded deterministically from the pair
(`seed`, document index), and all reductions are performed in a fixed order.
Two properties follow: the output is bit-identical for every value of
`n_threads`, which therefore affects wall time only, and the calibrated
p-values are reproducible under `seed`.

The study below evaluates VEM grids of ten models (five on the largest
corpus) on four synthetic corpus scales, from J = 200 documents with
W = 2,000 features up to J = 10,000 documents with W = 20,000 features,
timing the statistic pass, the bootstrap-calibrated pass
(B = `r bench$meta$n_boot`), and the goodness-of-fit evaluation for one,
two, and four threads, with the corresponding pure-R implementations as
baselines. The sweep ends at four threads
because that is the host's logical core count; thread settings beyond the
physical cores only measure oversubscription, so the benchmark driver caps
the sweep at the core count and extends it automatically on machines with
more cores. Peak memory is the maximum resident set size reported by the
operating system for a fresh R process per configuration. The study is
reproducible via `data-raw/benchmark-efficiency.R`; timings are
machine-dependent (recorded here: `r bench$meta$cpu`,
`r bench$meta$cores` logical cores, `r bench$meta$r_version`).

```{r bench-speed-table, echo = FALSE}
boot_wide <- dcast(bres[mode %in% c("boot", "boot_r")],
                   scale + J + W ~ config, value.var = "elapsed_sec")
setorder(boot_wide, J)
setcolorder(boot_wide, c("scale", "J", "W", "R bootstrap (0.11.0)"))
knitr::kable(boot_wide, digits = 1,
             caption = paste("Wall time (seconds) of the bootstrap-calibrated",
                             "pass across the grid, by thread count, with the",
                             "0.11.0 R implementation as baseline"))
```

```{r bench-speed-table-stat, echo = FALSE}
stat_wide <- dcast(bres[mode == "stat"],
                   scale + J + W ~ sprintf("%d thread%s", threads,
                                           ifelse(threads > 1, "s", "")),
                   value.var = "elapsed_sec")
setorder(stat_wide, J)
knitr::kable(stat_wide, digits = 2,
             caption = paste("Wall time (seconds) of the statistic pass",
                             "across the grid, by thread count"))
```

```{r bench-scaling-plot, echo = FALSE}
boot_runs <- bres[mode == "boot"]
boot_runs[, speedup := elapsed_sec[threads == 1] / elapsed_sec, by = scale]
baseline_r <- bres[mode == "boot_r", .(scale, base_r = elapsed_sec)]
boot_runs <- merge(boot_runs, baseline_r, by = "scale")
boot_runs[, vs_R := base_r / elapsed_sec]
best_large <- boot_runs[scale == "large"][which.min(elapsed_sec)]

ggplot(boot_runs, aes(x = threads, y = vs_R, color = scale)) +
  geom_line(linewidth = 0.7) +
  geom_point() +
  geom_hline(yintercept = 1, linetype = 3) +
  scale_x_continuous(breaks = sort(unique(boot_runs$threads))) +
  labs(x = "Threads", y = "Speedup over the 0.11.0 R bootstrap",
       color = "Corpus scale",
       title = "Speedup of the calibrated pass",
       subtitle = "Total speedup over the pre-0.12.0 R implementation, by thread count") +
  theme_bw()
```

```{r bench-ram, echo = FALSE}
ram <- bres[mode %in% c("boot", "boot_r")]
ram[, corpus := sprintf("J = %d, W = %d", J, W)]
ram[, corpus := factor(corpus, levels = unique(corpus[order(J)]))]
ram[, config := factor(config, levels = c("R bootstrap (0.11.0)",
                                          grep("C\\+\\+", unique(config),
                                               value = TRUE)))]

ggplot(ram, aes(x = corpus, y = max_rss_mb, fill = config)) +
  geom_col(position = "dodge") +
  labs(x = NULL, y = "Peak resident set size (MB)", fill = NULL,
       title = "Peak memory of the calibrated pass",
       subtitle = "Maximum resident set size of a fresh R process per configuration") +
  theme_bw() +
  theme(legend.position = "bottom")
```

Three regularities hold across the scales. First, the compiled bootstrap is
`r sprintf("%.1f", min(bres[mode == "boot_r", elapsed_sec] / bres[mode == "boot" & threads == 1, elapsed_sec]))`–`r sprintf("%.1f", max(bres[mode == "boot_r", elapsed_sec] / bres[mode == "boot" & threads == 1, elapsed_sec]))`×
faster than the R implementation before any threading, and the gain grows
with the vocabulary: the alias-table path costs $O(N_j)$ per replicate
where the bin-wise method costs $O(k_j)$, and wide corpora have
$k_j \gg N_j$. Second, the parallel
document loop scales almost linearly in the number of physical cores: the
best configuration on the largest corpus (`r best_large$threads` threads)
runs the calibrated pass `r sprintf("%.1f", best_large$vs_R)`× faster
than the 0.11.0 baseline, falling from
`r sprintf("%.1f", bres[scale == "large" & mode == "boot_r", elapsed_sec] / 60)` minutes
to `r sprintf("%.0f", best_large$elapsed_sec)` seconds.
The statistic pass shares the document-level parallelism and gains
`r sprintf("%.1f", bres[scale == "large" & mode == "stat" & threads == 1, elapsed_sec] / min(bres[scale == "large" & mode == "stat", elapsed_sec]))`×
at its best thread setting on the same corpus; its speedup is sub-linear
because the BLAS product remains serial. Third, peak
memory is governed by the objects the evaluation must hold, namely the
fitted models and, on the largest corpus, the exported envelope bins,
rather than by the sampling itself: the two implementations stay within
roughly 15% of one another in either direction, and the per-thread state
amounts to one result slot per document, so peak memory does not grow with
`n_threads`.

**The discrepancy indices.** The goodness-of-fit tools presented above
follow the same design. The harmonized partition of
`optop_make_partition()` keeps the running minimum across the model grid
one vocabulary block at a time in compiled code, so the full
J $\times$ W minimum matrix is never materialized, and the three
discrepancy indices are evaluated by a fused engine that computes all
requested metrics, for the model and for the no-topics baseline side, in a
single traversal of each document block. The evaluation consumes the
sparse counts directly, forms the baseline on the fly, and parallelizes
over documents under the contract stated above: results are bit-identical
for every value of `n_threads`, exposed on the index functions and on
`optop_make_partition()`. The table reports the wall time of the full
evaluation, the partition followed by a three-metric document-level
`optop_index_table()` over the grid, with the vectorized R implementation
that preceded the engine as baseline. The rare-word threshold is set so
that the harmonized support remains non-trivial at every scale of these
synthetic corpora; the computational cost of the evaluation does not
depend on that choice.

```{r bench-indices-table, echo = FALSE}
idx_wide <- dcast(bres[mode %in% c("indices", "indices_r")],
                  scale + J + W ~ config, value.var = "elapsed_sec")
setorder(idx_wide, J)
setnames(idx_wide, "R implementation (pre-engine)",
         "R base<br>(pre-engine)")
engine_cols <- grep("^engine, ", names(idx_wide), value = TRUE)
setnames(idx_wide, engine_cols,
         sub("^engine, ", "C++ base<br>", engine_cols))
setcolorder(idx_wide, c("scale", "J", "W",
                        "R base<br>(pre-engine)"))
knitr::kable(idx_wide, digits = 1,
             escape = FALSE,
             caption = paste("Wall time (seconds) of the goodness-of-fit",
                             "evaluation (partition plus three-metric index",
                             "table) by thread count, with the pre-engine R",
                             "implementation as baseline"))
```

```{r bench-indices-inline, include = FALSE}
idx_r_large <- bres[scale == "large" & mode == "indices_r", elapsed_sec]
idx_best_large <- bres[scale == "large" & mode == "indices"][which.min(elapsed_sec)]
idx_ram_r <- bres[scale == "large" & mode == "indices_r", max_rss_mb]
idx_ram_cpp <- bres[scale == "large" & mode == "indices" &
                      threads == idx_best_large$threads, max_rss_mb]
```

On the largest corpus the evaluation falls from
`r sprintf("%.1f", idx_r_large / 60)` minutes to
`r sprintf("%.1f", idx_best_large$elapsed_sec)` seconds at
`r idx_best_large$threads` threads, a
`r sprintf("%.1f", idx_r_large / idx_best_large$elapsed_sec)`-fold
reduction, and peak memory falls from
`r sprintf("%.1f", idx_ram_r / 1024)` GB to
`r sprintf("%.1f", idx_ram_cpp / 1024)` GB, since neither the minimum
matrix nor the dense count and baseline blocks are formed.

On toolchains without OpenMP support (notably the default macOS compiler)
the parallel loops run single-threaded; the thread-independent speedup of
the compiled bootstrap and of the index engine is retained, while the
scaling with `n_threads` is not. The internal helper
`OpTop:::optop_openmp_available()` reports whether the installed build
supports OpenMP.

## Notes

- The legacy stability helpers (`topic_stability()`, `agg_topic_stability()`,
  `agg_document_stability()`, `get_topic_models()`) are deprecated and will
  be removed in a future release; the discrepancy indices shown here are
  their replacement.
- `selection = "legacy"` reproduces the pre-0.9.9 computation exactly for
  back-compatibility and is also deprecated.
- The chi-square reference for Test 1 is asymptotic rather than exact; the
  calibrated p-values presented above (`calibrate = "bootstrap"` or
  `calibrate = "moment"`) address this, and `?optimal_topic` documents the
  underlying statistical reasoning.

## References

Haldane, J. B. S. (1937). The exact value of the moments of the distribution
of chi-square. *Biometrika*, 29, 133–143.

Lewis, C. M. and Grossetti, F. (2022). A statistical approach for optimal
topic model identification. *Journal of Machine Learning Research*, 23(58),
1–20. <https://jmlr.org/papers/v23/19-297.html>

Lewis, C. M. and Grossetti, F. (2026). Goodness-of-fit indices and
diagnostics for topic models. Working paper.
