---
title: "Selecting K with OpTop"
subtitle: "Parametric tests and discrepancy indices for LDA models"
author: "Francesco Grossetti"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Selecting K with OpTop}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<style>
/* the html_vignette theme styles section h4s with a left margin and a bottom
   border; the author/date lines are h4s too and would inherit both */
h4.author, h4.date { margin-left: 0; border-bottom: none; }
</style>

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.2,
  fig.align = "center"
)
library(OpTop)
library(data.table)
library(ggplot2)

# Precomputed results shipped with the package (a few KB): tables derived from
# the full fitted grid, plus the captured console output of the optimal_topic()
# calls below.
res <- readRDS(system.file("extdata", "optop-vignette-results.rds",
                           package = "OpTop"))
meta <- res$meta
idx_tab <- res$inaugural$index_table
word_r2 <- res$inaugural$word_snapshot
chi_cal_mm <- res$inaugural$chi_cal_mm
sim_chi <- res$sim$chi
sim_picks <- res$sim$picks
sim_idx_tab <- res$sim$index_table

# The fitted grid itself (~170 MB) cannot ship with the package and lives in
# data-raw/ on the maintainer's machine. When it is available, every
# optimal_topic() call below runs live; when it is not (e.g. R CMD check
# re-building the vignette from the pruned tarball), the same objects are
# assigned from the shipped results and the calls' captured console output is
# replayed, so the rendered document is the same either way.
models_path <- file.path("..", "data-raw", "VEM_models_inaugural.rds")
have_models <- file.exists(models_path)
if (have_models) {
  VEM_models <- readRDS(models_path)
  toks <- quanteda::data_corpus_inaugural |>
    quanteda::tokens(remove_punct = TRUE,
                     remove_symbols = TRUE,
                     remove_numbers = TRUE) |>
    quanteda::tokens_tolower() |>
    quanteda::tokens_remove(quanteda::stopwords())
  mydfm_sub <- quanteda::dfm_trim(quanteda::dfm(toks), min_termfreq = 5)
  weighted_dfm <- quanteda::dfm_weight(mydfm_sub, scheme = "prop")
} else {
  chi_seq <- res$inaugural$chi_by_q[["q0.95"]]
  chi_min <- data.table::copy(chi_seq)
  chi_by_q <- res$inaugural$chi_by_q
  chi_cal <- res$inaugural$chi_cal
}
```

OpTop implements the parametric tests of Lewis and Grossetti (2022, JMLR
23(58)) for identifying the optimal number of topics of a Latent Dirichlet
Allocation (LDA) model, together with a family of regression-style
goodness-of-fit indices. This vignette walks through both on a real corpus,
then closes with a simulation where the true number of topics is known.

The only step run offline is the LDA fitting itself
(`r length(meta$K_grid)` models); everything downstream — every
`optimal_topic()` call, table, plot, and number quoted in the text — is
computed from those fits.

## Setting up: corpus, grid, models

We use the U.S. Presidential Inaugural Address corpus that ships with
**quanteda** — the same corpus as the paper's case study — with a standard
preprocessing pipeline and a vocabulary trim. The grid spans
K = `r min(meta$K_grid)` to `r max(meta$K_grid)` in steps of 2.

```{r fitting, eval = FALSE}
library(OpTop)

ncores <- 6

# Tokenize
toks <- quanteda::data_corpus_inaugural |>
  quanteda::tokens(remove_punct = TRUE,
                   remove_symbols = TRUE,
                   remove_numbers = TRUE) |>
  quanteda::tokens_tolower() |>
  quanteda::tokens_remove(quanteda::stopwords())

# Create DFM (with raw counts as well as proportions)
mydfm        <- quanteda::dfm(toks)
mydfm_sub    <- quanteda::dfm_trim(mydfm, min_termfreq = 5)
weighted_dfm <- quanteda::dfm_weight(mydfm_sub, scheme = "prop")

K_grid <- seq(2, 200, by = 2)

# Estimate VEM models in parallel
VEM_models <- parallel::mclapply(
  K_grid,
  function(k) {
    message("Estimating LDA with K = ", k)
    topicmodels::LDA(x = mydfm_sub, k = k)
  },
  mc.cores = ncores
)
```

After trimming, the corpus has `r meta$ndoc` documents and
`r meta$nfeat_trimmed` features (down from `r meta$nfeat_full`). The VEM fits
above are unseeded, so re-running the pipeline can shift individual
statistics slightly; the numbers quoted in this vignette come from one such
run.

`optimal_topic()` consumes **word proportions** (`weighted_dfm` above), while
the discrepancy indices of the second part consume the raw **counts** — keep
both objects around. With `verbose = TRUE` the function reports its setup,
progress across the grid, the selected model with the rule that chose it, and
the elapsed time:

```{r optimal-call-seq, eval = have_models}
# Use the sequential approach
chi_seq <- optimal_topic(lda_models = VEM_models,
                         weighted_dfm = weighted_dfm,
                         selection = "sequential",
                         q = 0.95,
                         alpha = 0.01,
                         do_plot = FALSE)
```

```{r optimal-call-seq-replay, eval = !have_models, echo = FALSE}
cat(res$inaugural$console_seq, sep = "\n")
```

```{r optimal-call-min, eval = have_models}
# Use the global minimum approach
chi_min <- optimal_topic(lda_models = VEM_models,
                         weighted_dfm = weighted_dfm,
                         selection = "min",
                         q = 0.95,
                         alpha = 0.01,
                         do_plot = FALSE)
```

```{r optimal-call-min-replay, eval = !have_models, echo = FALSE}
cat(res$inaugural$console_min, sep = "\n")
```

## Reading the result

`optimal_topic()` returns a `data.table` with one row per model:

```{r table-head}
head(chi_seq)
```

- `OpTop` is the **standardized** Test 1 statistic (the raw chi-square of
  Equation 8 divided by its degrees of freedom) — the quantity plotted in the
  paper's Figure 2. Under a model that fits, it fluctuates around 1.
- `df` is `sum_j P_j`, the total count of "relatively important" words across
  documents.
- `pval` is the upper-tail p-value of the raw statistic: small values mean
  the observed discrepancy is larger than sampling noise can explain, i.e.
  the K-topic model is rejected.

The `selection` rule affects which K is *reported* as optimal (and the plot
guides), never the table itself — the two calls above return the same data:

```{r rule-independence}
all.equal(chi_seq, chi_min)
```

### The two selection rules

Both rules are one-liners on the returned table, so it is worth seeing them
spelled out:

```{r selection-rules}
alpha <- 0.01

# "sequential": the smallest K the test fails to reject
seq_pick <- chi_seq[pval > alpha, topic][1]

# "min": the K with the minimum standardized statistic
min_pick <- chi_seq[which.min(OpTop), topic]

c(sequential = seq_pick, min = min_pick)
```

**`selection = "sequential"`** implements the classical scan for model order:
walk K upward and stop at the first model the test *fails to reject*
(`pval > alpha`). The idea is parsimony — take the smallest K that the data
cannot distinguish from adequate. On this corpus the sequential rule selects
**K = `r seq_pick`**. (If every model were rejected, `optimal_topic()` falls
back to the global minimum with a warning.)

**`selection = "min"`** reproduces the published case study: take the K with
the minimum standardized statistic. Here that is **K = `r min_pick`** —
remarkably close to the 92 the paper reports on the untrimmed vocabulary of
the same corpus, and (as the q-sweep below shows) insensitive to the choice
of `q`. One practical remark: the minimum rule needs a grid that extends past
the flattening of the curve to settle — on a grid truncated at K = 100 the
minimum sits at the boundary, and only the full K ≤ 200 scan lets it come to
rest. The sequential rule, by construction, stops at the first adequate K and
does not care how far the grid extends.

A look at the p-value column explains when and why the two rules diverge:

```{r pval-summary}
summary(chi_seq$pval)
```

With `df` of the order of
`r format(round(median(chi_seq$df)), big.mark = ",")`, the chi-square
reference distribution is extremely concentrated, so p-values saturate:
models whose standardized statistic is materially above 1 get p-values of
essentially 0, and models at or below 1 get p-values of essentially 1. When
every model is rejected, the sequential rule falls back to the global minimum
(with a warning); when several are accepted it stops at the smallest — which
may sit well before the minimum. This is a documented calibration caveat of
the chi-square yardstick (see `?optimal_topic`): in saturated regimes the
*shape* of the `OpTop` curve, and its minimum, carry the information.

Because the table is the return value, any plot is one `ggplot2` call away:

```{r optop-curve}
ggplot(chi_seq, aes(x = topic, y = OpTop)) +
  geom_line(linewidth = 0.8, color = "royalblue") +
  geom_vline(xintercept = seq_pick, linetype = 2) +
  geom_vline(xintercept = min_pick, linetype = 3) +
  annotate("point", x = min_pick, y = min(chi_seq$OpTop),
           color = "red", shape = 4, size = 4) +
  labs(x = "Topics", y = expression(OpTop[J]^{K}),
       title = "Standardized Test 1 statistic across the grid",
       subtitle = sprintf("sequential pick: K = %d (dashed) | global minimum: K = %d (dotted)",
                          seq_pick, min_pick)) +
  theme_bw()
```

## Calibrated p-values

The saturation above has a precise cause (see the *Calibration* section of
`?optimal_topic`): the chi-square reference is a yardstick, not the exact
null law of the statistic. `optimal_topic()` can therefore *calibrate* the
p-values under the **fitted-model null** — "document `j` is
`Multinomial(N_j, I_j)` with the model's own fitted probabilities" — which is
simulated exactly and cheaply on the collapsed envelope bins. The only new
ingredient is the vector of document lengths:

```{r optimal-call-cal, eval = have_models}
doc_lens <- quanteda::ntoken(mydfm_sub)

chi_cal <- optimal_topic(lda_models = VEM_models,
                         weighted_dfm = weighted_dfm,
                         q = 0.95,
                         alpha = 0.01,
                         calibrate = "bootstrap",
                         n_boot = 200,
                         doc_lengths = doc_lens,
                         seed = 20260703,
                         do_plot = FALSE)
```

```{r optimal-call-cal-replay, eval = !have_models, echo = FALSE}
cat(res$inaugural$console_cal, sep = "\n")
```

The table now carries two p-value columns: `pval` (calibrated — what the
selection rules use) and `pval_chisq` (the asymptotic value, kept for
comparison):

```{r cal-table}
head(chi_cal)
```

```{r cal-plot}
pv <- melt(chi_cal[, .(topic, calibrated = pval, `chi-square` = pval_chisq)],
           id.vars = "topic", variable.name = "reference",
           value.name = "pvalue")

ggplot(pv, aes(x = topic, y = pvalue, color = reference)) +
  geom_line(linewidth = 0.7) +
  geom_hline(yintercept = alpha, linetype = 3) +
  labs(x = "Topics", y = "p-value",
       title = "Calibration replaces a saturated reference with a usable one",
       subtitle = sprintf("dotted line: alpha = %.2f", alpha)) +
  theme_bw()
```

```{r cal-picks}
c(sequential_chisq = chi_seq[pval > alpha, topic][1],
  sequential_calibrated = chi_cal[pval > alpha, topic][1])
```

Where the asymptotic p-values jump between 0 and 1, the calibrated series
moves through the interior, and `alpha` becomes a real Type-I error rate with
respect to the fitted-model null (at bootstrap resolution
`1 / (n_boot + 1)`). Note how much earlier the calibrated scan stops: once
the reference distribution has the right scale, the residual discrepancy at
small K is already within what sampling noise produces under the fitted
model, so the sequential rule reads "adequate" much sooner — while the
global-minimum rule and the discrepancy indices below remain the tools for
asking how much *more* fit additional topics buy. Two practical notes:
`calibrate = "moment"` gives a closed-form approximation of the same null
(no simulation — on this grid its sequential pick is
K = `r chi_cal_mm[pval > alpha, topic][1]`, agreeing with the bootstrap),
and both calibrations hold the estimated `theta` and `phi` fixed, so they do
not account for estimation noise in the LDA parameters themselves.

## The role of `q`

`q` controls the per-document envelope: sort each document's fitted word
probabilities in decreasing order and keep the smallest head whose cumulative
mass exceeds `q` — those are the `P_j` "relatively important" words; the rest
are collapsed into a single bin whose mass stays strictly below `1 - q`. In
the paper's notation `q = 1 - I^K`; its numerical setting `I^K = 0.05`
corresponds to the default `q = 0.95`.

Re-running the test across a sweep of `q` values is a direct sensitivity
check:

```{r q-sweep-run, eval = have_models}
q_grid <- c(0.80, 0.90, 0.95, 0.99)

chi_by_q <- lapply(q_grid, function(q) {
  optimal_topic(VEM_models, weighted_dfm, q = q, alpha = alpha,
                do_plot = FALSE)
})
names(chi_by_q) <- sprintf("q%.2f", q_grid)
```

```{r q-sweep-plot}
q_curves <- rbindlist(
  lapply(names(chi_by_q), function(nm) {
    data.table(q = as.numeric(sub("q", "", nm)), chi_by_q[[nm]])
  })
)

ggplot(q_curves, aes(x = topic, y = OpTop, color = factor(q))) +
  geom_line(linewidth = 0.7) +
  labs(x = "Topics", y = expression(OpTop[J]^{K}), color = "q",
       title = "The envelope parameter shifts the level, not the story") +
  theme_bw()
```

Applying the two selection rules to each element of the sweep — the same
one-liners as before — shows how the picks react to `q`:

```{r q-picks}
picks_by_q <- data.table(
  q = as.numeric(sub("q", "", names(chi_by_q))),
  sequential = sapply(chi_by_q, function(tab) tab[pval > alpha, topic][1]),
  min = sapply(chi_by_q, function(tab) tab[which.min(OpTop), topic])
)

knitr::kable(picks_by_q, caption = "Selected K by rule as q varies")
```

Larger `q` keeps more words per document, so the degrees of freedom grow and
the envelope covers more of each document's distribution; smaller `q` makes
the comparison coarser and cheaper. The sequential pick moves with `q` while
the global minimum stays put. The default `q = 0.95` follows the paper.

## Goodness-of-fit: the discrepancy indices

The second family of tools asks a complementary question: not *which K is
optimal* but *how much of the word-count structure a given model explains*.
The indices compare each model against the no-topics corpus baseline on a
**harmonized support**: a fixed, document-specific set of rare words
(computed across the whole grid) is collapsed so that every K is evaluated on
the same bins. They consume raw counts:

```{r indices-call, eval = FALSE}
dtm_counts <- methods::as(mydfm_sub, "CsparseMatrix")
partition <- optop_make_partition(VEM_models, dtm_counts, c = 5)
baseline <- optop_make_baseline(dtm_counts)

idx_tab <- optop_index_table(VEM_models, dtm_counts,
                             metrics = c("se", "chisq", "deviance"),
                             partition = partition, baseline = baseline,
                             macro = TRUE)
```

```{r indices-plot}
idx <- melt(idx_tab,
            id.vars = "K",
            measure.vars = c("R2_SE", "R2_chisq", "R2_dev"),
            variable.name = "metric", value.name = "R2")

ggplot(idx, aes(x = K, y = R2, color = metric)) +
  geom_line(linewidth = 0.7) +
  labs(x = "Topics", y = expression(R^2),
       title = "Discrepancy indices across the grid (document level, micro)") +
  theme_bw()
```

All three indices are regression-style R-squared analogues: 0 means the model
explains no more than the corpus-wide baseline, 1 means a perfect account of
the (harmonized) word counts. They typically rise steeply at small K and
flatten once additional topics stop buying explanatory power — the elbow
region is where the Test 1 machinery above places the optimum.

The indices also exist at the **word level**, which is useful diagnostics:
which words does the selected model capture well, and which does it miss?
For the sequential pick (K = `r word_r2$k`), the micro word-level chi-square
index is `r round(word_r2$micro, 3)`; the tails look like this:

```{r word-snapshot}
knitr::kable(
  data.frame(
    `best words` = names(word_r2$best),
    R2_best = round(unname(word_r2$best), 3),
    `worst words` = names(word_r2$worst),
    R2_worst = round(unname(word_r2$worst), 3),
    check.names = FALSE
  ),
  caption = sprintf("Word-level chi-square R2 at K = %d: the 10 best and worst captured words",
                    word_r2$k)
)
```

## A simulation with known truth

Real corpora never reveal their true K. To see the machinery work where the
answer is known, we simulate one corpus from the LDA generative process
itself: `r meta$J_sim` documents over `r meta$W_sim` words drawn from
**`r meta$true_K` topics**, using `sim_dfm()` (a thin wrapper around
`LDATS::sim_LDA_data()`), then re-fit a grid around the truth.

```{r sim-call, eval = FALSE}
rdirichlet <- function(n, alpha) {
  g <- matrix(rgamma(n * length(alpha), shape = alpha, rate = 1),
              nrow = n, byrow = TRUE)
  g / rowSums(g)
}

set.seed(20260703)
true_K <- 10L
Theta_true <- rdirichlet(100L, rep(0.3, true_K))    # documents x topics
Beta_true <- rdirichlet(true_K, rep(0.05, 1500L))   # topics x words
doc_len <- sample(500:1500, 100L, replace = TRUE)

sim_corpus <- sim_dfm(DTW = Theta_true, TWW = Beta_true,
                      doc_length = doc_len, seed = 42)

sim_models <- parallel::mclapply(
  seq(2, 20, by = 2),
  function(k) topicmodels::LDA(sim_corpus, k = k,
                               control = list(seed = 2000 + k)),
  mc.cores = ncores
)

sim_chi <- optimal_topic(sim_models,
                         quanteda::dfm_weight(sim_corpus, scheme = "prop"),
                         q = 0.95, alpha = alpha, verbose = TRUE)
```

```{r sim-plot}
ggplot(sim_chi, aes(x = topic, y = OpTop)) +
  geom_line(linewidth = 0.8, color = "royalblue") +
  geom_vline(xintercept = meta$true_K, color = "red", linetype = 1) +
  geom_vline(xintercept = sim_picks$sequential, linetype = 2) +
  geom_vline(xintercept = sim_picks$min, linetype = 3) +
  labs(x = "Topics", y = expression(OpTop[J]^{K}),
       title = "Simulated corpus: the truth is K = 10 (solid red)",
       subtitle = sprintf("sequential pick: K = %d (dashed) | global minimum: K = %d (dotted)",
                          sim_picks$sequential, sim_picks$min)) +
  theme_bw()
```

```{r sim-indices}
sim_idx <- melt(sim_idx_tab,
                id.vars = "K",
                measure.vars = c("R2_SE", "R2_chisq", "R2_dev"),
                variable.name = "metric", value.name = "R2")

ggplot(sim_idx, aes(x = K, y = R2, color = metric)) +
  geom_line(linewidth = 0.7) +
  geom_vline(xintercept = meta$true_K, color = "red", linetype = 1) +
  labs(x = "Topics", y = expression(R^2),
       title = "Discrepancy indices on the simulated corpus") +
  theme_bw()
```

The statistic bends in the neighborhood of the true K and the indices stop
improving there — the same qualitative behavior the paper's simulation study
reports (its Figure 7). As the paper notes, the identified optimum is not
always exactly `K_sim` but typically adjacent to it: here the sequential rule
stops at K = `r sim_picks$sequential` with the truth at `r meta$true_K`,
while the global minimum drifts to the grid edge as the curve keeps creeping
down past the bend. Treat the tools as locating the *region* of adequate K,
with the discrepancy indices quantifying what each candidate buys you.

## Notes

- The legacy stability helpers (`topic_stability()`, `agg_topic_stability()`,
  `agg_document_stability()`, `get_topic_models()`) are deprecated and will
  be removed in a future release; the discrepancy indices shown here are
  their replacement.
- `selection = "legacy"` reproduces the pre-0.9.9 computation exactly for
  back-compatibility and is also deprecated.
- The chi-square reference for Test 1 is a yardstick, not an exact null law;
  a calibration layer (parametric-bootstrap null) is on the roadmap — see the
  package's `AUDIT.md` for the technical background.

## References

Lewis, C. M. and Grossetti, F. (2022). A statistical approach for optimal
topic model identification. *Journal of Machine Learning Research*, 23(58),
1–20. <https://jmlr.org/papers/v23/19-297.html>
