---
title: "OpTop at very large scale"
output: rmarkdown::html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
```

This article is the operations manual for running OpTop on corpora that
outgrow a single sparse matrix: millions to tens of millions of documents,
vocabularies in the tens of thousands, evaluated on server machines. Nothing
here changes the statistics. Every tool below computes exactly what the
in-memory workflow computes, and for ordinary corpora (up to a few million
documents that fit comfortably in memory) you do not need any of it: pass
your dfm or dgCMatrix directly and stop reading.

The article covers, in order: why a single matrix eventually fails, sharding
with `optop_corpus()`, lazy model loaders and the cache budget, the sharded
end-to-end pipeline, thread and BLAS configuration, and memory planning with
reference timings.

## Why a single matrix eventually fails

R's sparse matrix container (`dgCMatrix`) stores nonzero positions in a
32-bit index, so one matrix holds at most $2^{31} - 1$ nonzero cells. An
accounting corpus of 14 million filings with a few hundred distinct terms
each crosses that bound. Memory follows close behind: the counts, their
transpose, and the per-model fitted quantities all scale with the token
count.

OpTop's compiled kernels have no shape limit of their own (the boundary is
zero-copy over the sparse slots, and all index arithmetic is 64-bit). The
container is the bound, and sharding removes it.

## Sharding with optop_corpus()

`optop_corpus()` wraps an ordered collection of document shards, each a
matrix small enough for the container. Three forms are accepted:

```{r corpus-forms}
library(OpTop)

# 1) a single matrix: a corpus with one shard, exactly equivalent to
#    passing the matrix itself
corpus <- optop_corpus(dtm)

# 2) an in-memory list of shards: documents split by rows, all shards
#    sharing the same vocabulary in the same order
corpus <- optop_corpus(list(dtm_2015, dtm_2016, dtm_2017))

# 3) file-backed shards with a reader: nothing is loaded until a shard is
#    evaluated, and only one shard is in memory at a time
corpus <- optop_corpus(
  sprintf("shards/counts-%02d.qs2", 1:24),
  reader = function(path) qs2::qs_read(path)
)
```

Shards must carry unique, globally disjoint document names and identical
column names. The constructor validates what it can see immediately and
defers file-backed validation to first use.

The exactness contract is worth stating precisely. Per-document quantities
(statistic terms, discrepancies, calibration draws) are bit-identical to an
unsharded evaluation: every document owns a deterministic random stream
keyed by its global position, and per-document arithmetic does not depend on
which shard holds the document. Corpus-level sums group floating-point
additions by shard, so totals agree with the unsharded run to summation
order, a few units in the last place. A one-shard corpus reproduces the
plain call bit for bit, calibrated p-values included.

## Lazy model loaders and the cache budget

A grid of fitted models for a large corpus can itself exhaust memory: theta
alone is $J \times K$ doubles per model. Wherever a list of models is
accepted, elements may be loader functions returning a fit; each is
materialized on demand and released:

```{r loaders}
models <- lapply(sprintf("fits/lda-k%02d.qs2", seq(10, 80, by = 10)),
                 function(path) {
                   force(path)
                   function() qs2::qs_read(path)
                 })

test1 <- optimal_topic(models, corpus,
                       doc_lengths = doc_lengths,  # named, see below
                       n_threads = 16)
```

Extracted weight matrices are cached across shards while the whole grid
fits a budget, 256 MB by default. On a large server, raise it so the grid
stays resident across shard passes:

```{r cache}
options(optop.cache_mb = 4096)
```

Results never depend on the budget; only wall time does.

## The sharded pipeline end to end

Every stage accepts the corpus where it accepts a matrix. Two details
change relative to the in-memory workflow: `doc_lengths` must be a named
vector when calibrating (row order across shards is not a reliable
identity at this scale), and the counts corpus and the weighted corpus
must shard the documents identically.

```{r pipeline}
# Test 1 with bootstrap calibration
doc_lengths <- unlist(lapply(shard_paths, function(p) {
  m <- qs2::qs_read(p)
  stats::setNames(Matrix::rowSums(m), rownames(m))
}))

wprop <- optop_corpus(weighted_paths, reader = function(p) qs2::qs_read(p))
test1 <- optimal_topic(models, wprop,
                       calibrate = "bootstrap", n_boot = 200,
                       doc_lengths = doc_lengths, seed = 42,
                       n_threads = 16)

# goodness-of-fit indices on the counts corpus
counts <- optop_corpus(shard_paths, reader = function(p) qs2::qs_read(p))
partition <- optop_make_partition(models, counts, c = 1, n_threads = 16)
baseline <- optop_make_baseline(counts)
tab <- optop_index_table(models, counts,
                         metrics = c("se", "chisq", "deviance"),
                         macro = TRUE,
                         partition = partition, baseline = baseline,
                         n_threads = 16)
```

Held-out evaluation (`optop_index_holdout()`, `optop_moment_test()`)
accepts an `optop_corpus` for the evaluation documents as well; the
training-side objects (baseline, instruments) are ordinary vectors and
never large.

All long-running kernels check for user interrupts between work chunks, so
a mistaken launch on a 14-million-document corpus can be cancelled from the
console rather than killed.

## Threads and BLAS

`n_threads` controls the OpenMP parallelism of the compiled kernels. The
contract is strict: results are bit-identical for any value, because every
document owns its result slot and all reductions run in a fixed order.
Choose it freely; on builds without OpenMP (the default macOS toolchain)
the value is ignored and the kernels run single-threaded.

Two thread pools can interact, and one configuration deserves care. The
per-block fitted-probability product runs through your R installation's
BLAS. Inside a single OpTop call there is no nesting: the BLAS product runs
in serial code, outside the OpenMP regions. But a common server pattern
wraps model fitting or per-period evaluations in `parallel::mclapply()`,
and if the linked BLAS is itself multithreaded (OpenBLAS, MKL, Apple
Accelerate), every forked worker spawns its own BLAS pool and the machine
oversubscribes. Cap the BLAS pool in that pattern:

```{r blas}
# either an environment variable before starting R
#   OPENBLAS_NUM_THREADS=1  (or MKL_NUM_THREADS, VECLIB_MAXIMUM_THREADS)
# or at runtime
RhpcBLASctl::blas_set_num_threads(1)
```

As a rule: give threads either to OpTop (`n_threads`) or to the outer
`mclapply()`, not both, and keep the BLAS pool at one thread whenever the
outer level forks.

## Memory planning and reference timings

The partition is the object to size in advance. It stores the complement
of the rare set, bounded by one entry per `c` tokens: at one billion tokens
and `c = 1` that is at most $10^9$ 32-bit word indices, about 4 GB, and
proportionally less for larger `c`. No dense $J \times W$ object of any
kind is materialized anywhere; the pre-0.15.0 dense mask at the largest
benchmark size below would have needed 200 GB.

Reference timings from the development benchmark (synthetic Zipf corpora,
two toy models per grid, `dev/benchmark-scale.md` in the repository; see
the NEWS entries of 0.15.0 and 0.17.0 for the architecture behind them):

|     J|     W|      nnz| partition_s| partition_mb| doc_index_s| optimal_topic_s|
|-----:|-----:|--------:|-----------:|------------:|-----------:|---------------:|
| 1e+05| 30000|  4928396|         0.1|         51.2|         0.1|            40.0|
| 5e+05| 50000| 24990158|         0.6|        226.4|         0.4|           357.2|
| 1e+06| 50000| 42378862|         0.9|        400.3|         0.8|           739.4|

The `optimal_topic` column is dominated by the per-model BLAS product over
the full vocabulary, which is inherent to the statistic; the partition and
the discrepancy indices run in seconds at a million documents because they
scale with the token count, not with $J \times W$.
